# Day 1

## Welcome + Mandatory Formalities

- [About me](https://rmshksu.github.io)

- [Course website](https://rmshksu.github.io/stat341-spring26/about/)

  - Canvas
  
  - Email, don't Canvas message

- [Syllabus](https://rmshksu.github.io/stat341-spring26/misce/Syllabus.epub)

- What's this course about?
  
- Is it right for you?
  
- Statistical programming
  
- Reproducibility and academic honesty
  
- Since I only recognize 32 of you...
  
    - How I teach
  
    - Grades


```{r , message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
library(kableExtra)
library(tidyverse)

card_kable <- function(
  data,
  title    = NULL,
  subtitle = NULL,
  caption  = NULL,
  width    = "85%",
  header_color = "#51288590"
) {
  # Build the kable table (HTML)
  tbl <- knitr::kable(
    data,
    format = "html",
    escape = FALSE
  ) |>
    kableExtra::kable_styling(
      full_width = FALSE,
      bootstrap_options = c("condensed", "hover")
    )

  tbl_html <- as.character(tbl)

  # Title bar
  header_html <- if (!is.null(title)) {
    sprintf(
      '<div style="background:%s; color:white; padding:6px 10px; font-weight:600; 
                  border-radius:8px 8px 0 0;">%s</div>',
      header_color, title
    )
  } else {
    ""
  }

  # Subtitle text (optional)
  subtitle_html <- if (!is.null(subtitle)) {
    sprintf(
      '<div style="padding:6px 10px; font-size:0.9em; color:#555;">%s</div>',
      subtitle
    )
  } else {
    ""
  }

  # Caption text (optional)
  caption_html <- if (!is.null(caption)) {
    sprintf(
      '<div style="padding:6px 10px; font-size:0.8em; color:#777; border-top:1px solid #eee;">%s</div>',
      caption
    )
  } else {
    ""
  }

  card_html <- paste0(
    '<div style="border:1px solid #ddd; border-radius:8px; margin:1em auto; width:', width, 
    '; box-shadow:0 2px 5px rgba(0,0,0,0.06); background:white;">',
      header_html,
      subtitle_html,
      '<div style="padding:8px 10px;">',
        tbl_html,
      '</div>',
      caption_html,
    '</div>'
  )

  knitr::asis_output(card_html)
}
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
data = data.frame(Week = 1:17,
                  Topics = c("Review and math overview",
                               "Math and R programming",
                               "Statistical modeling",
                               "Linear models I",
                               "Linear models II",
                               "Hypothesis testing",
                               "Prediction",
                               "Inference",
                               "SPRING BREAK",
                               "Model checking",
                               "Review and catch-up",
                               "Assumptions and errors",
                               "Generalized linear models",
                               "Mixed Models",
                               "Class choice I",
                               "Class choice II",
                               "Finals"),
                  Assignments = c("HW 1 / Journal 1",
                                  "Journal 2",
                                  "HW 2 / Journal 3",
                                  "Journal 4",
                                  "HW 3 / Journal 5",
                                  "Journal 6",
                                  "HW 4 / Journal 7",
                                  "Project proposal / Journal 8",
                                  "",
                                  "Literature review",
                                  "Exploratory analysis",
                                  "Model descriptions",
                                  "Model fitting/checking",
                                  "Results and inference",
                                  "Rough draft",
                                  "Presentation",
                                  "Final Project + Report"))

card_kable(
  data    = data,
  title   = "Content Schedule",
  subtitle= "All dates are subject to change.",
  width   = "85%",
  header_color = "#51288590"
)
```

- About you!

```{r,echo=FALSE}
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/students.csv"
students = read.csv(url)
majors=as.data.frame(table(students$degreeProgram))
levels=as.data.frame(table(students$classLevel))
levels=data.frame(Var1=c("Sophomore","Junior","Senior"),
                  Freq=c(6,24,26))
par(mar = c(5, 10, 4, 2))
barplot(majors$Freq,
        names.arg = majors$Var1,
        horiz = TRUE,
        las = 1,
        xlim=c(0,36),
        cex.names = 0.8,
        col = "#51288899",
        main = "Degree Programs",
        xlab = "Count")
```

```{r,echo=FALSE}
par(mar = c(3, 4, 3, 1))
barplot(levels$Freq,
        names.arg = levels$Var1,
        las = 1,
        cex.names = 0.8,
        col = "#51288899",
        main = "Class Levels",
        ylab = "Count",
        ylim=c(0,30))
```

## Texts and Literature

- There's no required textbook for the course

    - There are *heavily* recommended texts however

<br>

>Faraway, J. J. (2025). *Linear models with R (3rd ed.).* Chapman and Hall/CRC.

>Weisberg, S. (2014). *Applied linear regression* (4th ed.). Wiley.

>Fieller, N. (2016). *Basics of matrix algebra for statistics with R.* CRC Press.

>Wickham, H., & Grolemund, G. (2017). *R for data science: Import, tidy, transform, visualize, and model data.* O’Reilly Media.

<br>

- There will be the occasional required reading

    - I'll always provide those passages
    
- Additional reading and resources will be made available over time

<br>

## Homeworks

- Project-based education

- Group work/collaboration

- Structure of homeworks:

  - Due every other week*
  
  - (2 person) Group assignments*
  
  - Different partners each time
  
*Except this week
  
- Purpose of homeworks:

  - Build comfort with R programming
  
  - Practice essential methods
  
  - "Interview" potential project partners
  
<br>
  
## Journals

- Weekly writing assignment

- 2-3 sentences answering each of the following:

1. What's something you've learned this week?

2. What's something you're struggling with?

3. Is there anything else you'd like me to know?
  
<br>

- These are "attendance"

    - Graded on based on completion

    - You can write about both classes that week
    
    - You can write about just one
    
- Getting in the habit of writing

- Personal reflection and engagement with material

- Honor system

<br>
  
## Final Project

- 2-3 person groups 

  - Working alone ([softly] Don't.)
  
- Propose a project (Week 8)

  - Academic research
  
  - Industry solutions
  
  - General interest
  
- Final presentation (Dead & Finals week)
  
- Final paper (Friday of finals)

<br>

## Checkpoint Assignments

- Starting after spring break

- Guide you through the process of building a project

- Graded on a 0%, 50%, 100% scale

    - Didn't do it, didn't try, or did it
    
- Don't let these overwhelm you

    - Don't slack either
    
<br>

## Review

### General Intro Stats

- Everyone *should* know these

- Regardless of *who* your intro stats instructor was

<br>

- Mean and standard deviation

\[
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i
\qquad\qquad
s = \sqrt{s^2} = \sqrt{\frac{1}{n-1} \sum_{i=1}(x_i - \bar{x})^2}
\]


```{r}
# base R dataset on fiji earthquakes
quakes = datasets::quakes

# mean magnitude
mean(quakes$mag)

# variance of magnitude
var(quakes$mag)

# standard deviation of magnitude
sd(quakes$mag)
```

- Quantiles/Quartiles

\[
\begin{array}{|c|c|c|c|c|}
\hline
\text{Min} & \text{Q}_1 & \text{Median} & \text{Q}_3 & \text{Max} \\
\hline
0^{th} & 25^{th} & 50^{th} & 75^{th} & 100^{th} \\
\hline
\end{array}
\]

```{r}
# five number summary as quantiles
quantile(quakes$mag, c(0,0.25,0.5,0.75,1))

# five number summary base R function
fivenum(quakes$mag)
```

- Histograms and Boxplots

```{r}
# base R histogram
hist(quakes$mag,
     xlab = "Magnitude", # x axis label
     main = "Earthquakes off Fiji", # title
     col = "white") # bar color
```

```{r}
hist(quakes$mag, # data
     xlab = "Magnitude",
     main = "Earthquakes off Fiji",
     col = "white")
abline(v = mean(quakes$mag), # vertical mean line
       lty = 2, # line type (dashed)
       lwd = 2, # line width
       col = "gold") # line color
```

```{r}
# margin control
par(mar = c(3,1,1,1)) # bottom, left, top, right
# base R boxplot
boxplot(quakes$mag,
        horizontal = T, # set as horizontal (T = TRUE)
        col = "white") # central box color
```

```{r}
par(mar = c(3,1,1,1))
boxplot(quakes$mag,
        horizontal = T,
        col = "white")
abline(v = c(4,4.3,4.6,4.9,5.7), # vertical quantile lines
       lty = 2,
       lwd = 2,
       # color palette
       col = c("#FED976","#FEB24C","#FD8D3C","#E31A1C","#800026"))
text(c(4.1,4.24,4.5,4.82,5.5), # x position
     c(0.9,1.15,0.9,1.15,0.9), # y position
     c("Min","Q1","Med","Q3","Max"), # quantile labels
     col = c("#FED976","#FEB24C","#FD8D3C","#E31A1C","#800026"))
```

- $z$-scores

\[
z = \frac{x - \bar{x}}{s}
\]

```{r}
# transform the entire dataset into z-scores
z = (quakes$mag - mean(quakes$mag))/sd(quakes$mag)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
par(mfrow = c(1,2), mar = c(4.1,4,3.8,1))
hist(quakes$mag,
     xlab = "Magnitude",
     main = "Earthquakes off Fiji",
     col = "white")
hist(z,
     xlab = "Standard Deviations",
     main = "Earthquake z-scores",
     col = "white")
```

- Normal distributions

\[
X \sim N(\mu,\sigma^2)
\qquad\qquad
\frac{X - \mu}{\sigma} = Z \sim N(0,1)
\]

```{r}
par(mar = c(4.5,4.5,1,1))
curve(dnorm(x), # standard normal density
      xlim = c(-4,4), # x axis limits
      lwd = 2,
      xlab = "X",
      ylab = "P(X=x)") # y axis label
```

```{r}
# z score for a 5.5 magnitude earthquake
z_score = (5.5 - mean(quakes$mag))/sd(quakes$mag)
1 - pnorm(z_score) # 1 - z probability (same as z-table)
```

- Scatterplots

```{r}
# base R scatterplot
plot(quakes$stations,quakes$mag, # x then y 
     xlab = "Number of stations reporting", # x axis label
     ylab = "Magnitude", # y axis label
     main = "Earthquakes off Fiji", # title
     pch = 20, # point shape
     col = "#00000080") # point color
```

- Least squares regression

\[
y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

```{r}
# predictor
x = quakes$stations

# response
y = quakes$mag

# base R linear model
m = lm(y ~ x)
coef(m) # coefficients
```

```{r}
plot(quakes$stations,quakes$mag,
     xlab = "Number of stations reporting",
     ylab = "Magnitude",
     main = "Earthquakes off Fiji",
     pch = 20,
     col = "#00000080")
# add regression line
lines(sort(quakes$stations),sort(m$fitted.values),
      lwd = 2,
      col = "gold") # line color
```

- Confidence intervals

```{r}
# 95% confidence interval

# mean magnitude
xbar_mag = mean(quakes$mag)
# standard deviation of magnitude
s_mag = sd(quakes$mag)
# sample size
n_mag = as.numeric(nrow(quakes))
# margin of error
moe_mag = 1.96*(s_mag/sqrt(n_mag))

xbar_mag - moe_mag # lower bound
xbar_mag # point estimate
xbar_mag + moe_mag # upper bound
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
hist(quakes$mag,
     xlab = "Magnitude",
     main = "Earthquakes off Fiji",
     col = "white")
abline(v = c(xbar_mag - moe_mag, xbar_mag, xbar_mag + moe_mag),
       lty = 2,
       lwd = 2,
       col = c("#FEB24C","#FD8D3C","#E31A1C"))
```

- Hypothesis tests

```{r message=FALSE, warning=FALSE}
# schizophrenic reaction time data
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/4c51b2017bb69c11d7ee918cac7d594c07b39549/schiz.txt"
schiz = data.frame(time = scan(url,skip=5),
                   schizophrenia = c(rep("no",30*11),rep("yes",30*6)))

library(dplyr) # R data wrangling package
sstats = schiz %>% 
  # group by yes/no schiz
  group_by(schizophrenia) %>% 
  # mean/var/sample size for yes/no strata
  summarise(xbar = mean(time),
            s2 = var(time),
            n = n())
# print dataframe
sstats
```

\[
\begin{aligned}
\text{H}_0: \mu_1 - \mu_2 = 0 \\
\\
\text{H}_a: \mu_1 - \mu_2 < 0
\end{aligned}
\]

<br>

\[
t^* = \frac{(\bar{x}_1 - \bar{x}_2) + (\mu_1 - \mu_2)}{\sqrt{(s_1^2/n_1) + (s_2^2/n_2)}}
\]

```{r}
# test statistics for independent difference in means
tstar = as.numeric((sstats[1,2] - sstats[2,2])/
                     sqrt((sstats[1,3]/sstats[1,4]) +
                            (sstats[2,3]/sstats[2,4])))
tstar

# p-vale for t-pivot test statistic
pt(tstar,min(sstats[,4])-1)
```

<br>

### Biometrics I

- Review for my students

- Potentially everyone

- It's okay to not know these

  - Come to office hours though

<br>

- Discrete expected value

\[
\mathbb{E}X = \sum_x x P(X=x)
\]

- Binomial distribution

  - Discrete, binary processes
  
  - Success/failure

\[
X \sim \text{Bin}(n,p)
\qquad \qquad 
\mathbb{E}X = np
\qquad \qquad 
\mathbb{V}X = np(1-p)
\]

- Poisson distribution

  - Positive discrete counts
  
\[
X \sim \text{Pois}(\lambda)
\qquad \qquad 
\mathbb{E}X = \mathbb{V}X = \lambda
\]

- Geometric distribution

  - Number of trials to reach success
  
\[
X \sim \text{Geom}(p) 
\qquad \qquad
\mathbb{E}X = 1/p
\]

- Hypergeometric distribution

  - Geometric distribution
  
  - Sampling without replacement
  
\[
X \sim \text{Hyper}(N,K,n) 
\qquad \qquad
\mathbb{E}X = n \frac{K}{N}
\]

<br>

- Continuous expected values

\[
\mathbb{E}X = \int_{\chi}x f_x(x)
\]

- Gamma distribution

  - Positive continuous values

\[
X \sim \text{Gamma}(\alpha,\theta)
\qquad \qquad 
\mathbb{E}X = \alpha \theta
\qquad \qquad 
\mathbb{V}X = \alpha \theta^2
\]

- Beta distribution

  - Proportions/probabilities

\[
X \sim \text{Beta}(\alpha,\beta)
\qquad \qquad 
\mathbb{E}X = \frac{\alpha}{\alpha + \beta}
\qquad \qquad 
\mathbb{V}X = \frac{\alpha \beta}{(\alpha + \beta)^2 (\alpha + \beta + 1)}
\]

- Uniform distribution

  - Equi-probable events
  
  - Defined boundaries (a and b)
  
\[
X \sim \text{Unif}(a,b)
\qquad \qquad 
\mathbb{E}X = \frac{1}{2}(b-a)
\qquad \qquad 
\mathbb{V}X = \frac{1}{12}(b-a)^2
\]

<br>

## Assignment 1

## Go away
