# Day 5

## Announcements

- Office hours for this week:

    - Wednesday: 9:00 AM - 10:59 AM
    
    - Thursday: 11:30 AM - 12:30 PM
    
    - Friday: 9:00 AM - 10:59 AM
    
- "Open door policy"

- By-appointment is always available

<br>

Journals

- 50/50 split on math/R

    - Learning 2 languages at once
    
    - We're done with the "programming" part
    
    - Homework 2 is by far the most intensive one for R
    
    - If you can do 100% of it you're set for the semester
    
- Re: Homeworks/missed journals/grades

- Purpose of the lecture notes/code blocks

<br>

## Review

- R Markdown

  - "Graves" " ` "
  
  - {r}, {r, echo=FALSE, message=FALSE, warning=FALSE}
  
  - {fig.align='center'}
  
  - HTML vs PDF
  
  - Headers
  
<br>
  
```
# First level

## Second level

### Third level

- Bullet

* Bullet

+ Bullet

1. Numbered bullet

a. Letter bullet

> Quote block
```

- Quarto vs. Rmd

<br>

- LaTeX

  - `$` vs `$$` vs `\[`
  
  - `\beta`, `\gamma`
  
  - `\bar{}`, `\hat{}`
  
  - `\begin{array}`, `\begin{bmatrix}`, `\begin{aligned}`
  
  - `\vspace{3mm}`, `\newpage`

<br>

**Any questions?**

<br>

## Statistical Modeling

- What is a model?

  - Scientific vs. Thought vs. Mathematical
  
  - Statistical?
  
- Why do we use models?

<br>

- Trend analysis

  - What's happening with total kjeldahl nitrogen in Kansas?

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(gamair)
library(nlme)
library(mgcv)
library(tidyverse)
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/cruisedata.csv"
cruise = read.csv(url, stringsAsFactors = FALSE)
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/kslr_tkn.csv"
kslr_tkn = read.csv(url, stringsAsFactors = FALSE)
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/framingham_heart_study.csv"
framingham = read.csv(url, stringsAsFactors = FALSE)
```

```{r,fig.align='center'}
plot(kslr_tkn$Year, kslr_tkn$UOM, 
     col="#00000060",
     xlab = "Year",
     ylab = "TKN (mg/L)")
```

- Simple linear regression

\[
y_i = \beta_0 + \beta_1 t_i + \epsilon_i
\]

\[
\epsilon_i \sim N(0, \sigma^2 \textbf{I})
\]

<br>

- Another way to write it:

\[
y_i \sim N(\mu,\sigma^2 \textbf{I})
\]

\[
\mu = \beta_0 + \beta_1 t_i
\]

```{r}
mod_df = data.frame(UOM = kslr_tkn$UOM,
                    Year = kslr_tkn$Year,
                    huc08 = kslr_tkn$huc08)

m1 = lm(UOM ~ Year, data = mod_df)
summary(m1)
```

```{r,fig.align='center'}
or = order(mod_df$Year)
plot(mod_df$Year, mod_df$UOM, 
     col="#D1D1D180",
     xlab = "Year",
     ylab = "TKN (mg/L)")
lines(mod_df$Year[or], predict(m1, type="response")[or], lwd = 2, col = "#512885")
```

<br>

- What about space?

  - Add a "cluster effect"
  
\[
y_{ij} = \beta_0 + \beta_1 t_i + \beta_2 s_j \epsilon_{ij}
\]

\[
\epsilon_{ij} \sim N(0, \sigma^2 \textbf{I})
\]

- $s_j$ pertains to the "HUC" levels

  - Think of it as a categorical predictor
  
  - Goes to $0$ based on which HUC is present

```{r}
m2 = lm(UOM ~ Year + as.factor(huc08), data = mod_df)
summary(m2)
```

```{r, echo=FALSE,warning=FALSE,message=FALSE,fig.align='center'}
xg = seq(min(mod_df$Year, na.rm = TRUE),
         max(mod_df$Year, na.rm = TRUE),
         length.out = 200)
levs = levels(model.frame(m2)[["as.factor(huc08)"]])
cols = seq_along(levs)

plot(mod_df$Year, mod_df$UOM, 
     col="#D1D1D180",
     xlab = "Year",
     ylab = "TKN (mg/L)")
for (i in 1:length(levs)) {
  lv = levs[i]
  nd = data.frame(Year = xg, huc08 = factor(lv, levels = levs))
  lines(xg, predict(m2, nd), lwd = 1, col = cols[i])
}
legend("topright", legend = levs, lwd = 1, col = cols, bty = "n")
```

- Where are we headed?

<br>

## Stringing sentences

- Correlative inference

  - Which cruise liner should you avoid?

```{r cruisechunk1}
par(mar = c(5, 9, 4, 2))
boxplot(cruise$infected_proportion ~ cruise$liner,
        xlab = "Proportion infected",
        ylab = "",
        main = "Norovirus outbreaks on cruise liners",
        horizontal = TRUE,
        las = 1,
        cex.axis = 0.7,
        col = "#D1D1D190")
```

\[
\left[y|\alpha, \beta \right] \equiv \text{Beta}(\alpha,\beta)
\]

\[
\alpha = \mu \phi \qquad \beta = (1 - \mu) \phi \qquad g(\mu) = \textbf{X}^\prime \boldsymbol{\beta}
\]

```{r cruisechunk2}
m1 = gam(infected_proportion ~ as.factor(liner) - 1,
         data = cruise, family = betar(link = "logit"))
```

```{r cruisechunk3, echo=FALSE,warning=FALSE,message=FALSE,fig.align='center'}
# liner levels exactly as used in the fitted model
levs <- levels(model.frame(m1)[["as.factor(liner)"]])

# prediction data: one row per liner
nd <- data.frame(liner = factor(levs, levels = levs))

# get fitted linear predictor + SE per liner
pr <- predict(m1, newdata = nd, type = "link", se.fit = TRUE)

# 95% CI on link scale, then transform to response
invlogit <- m1$family$linkinv
fit_r <- invlogit(pr$fit)
lo_r  <- invlogit(pr$fit - 1.96 * pr$se.fit)
hi_r  <- invlogit(pr$fit + 1.96 * pr$se.fit)

# plot CI segments only, labeled by the original liner names
op <- par(mar = c(5, 10, 4, 2))

plot(
  NULL,
  xlim = range(c(lo_r, hi_r)),
  ylim = c(0.5, length(levs) + 0.5),
  yaxt = "n",
  xlab = "Predicted infected proportion (95% CI)",
  ylab = "",
  main = "Predicted mean infection by liner"
)

segments(lo_r, seq_along(levs), hi_r, seq_along(levs), lwd = 3)

axis(2, at = seq_along(levs), labels = levs, las = 1, cex.axis = 0.75)

par(op)
```

- Seems like Fred Olsen is a bad idea...

```{r cruisechunk4, echo=FALSE,warning=FALSE,message=FALSE,fig.align='center'}
freq_df = as.data.frame(table(cruise$liner))
freq_df = data.frame("Liner" = freq_df$Var1[1:10],
                     "Frequency" = freq_df$Freq[1:10])
card_kable(data = freq_df,
           title = "Frequency distribution of cruise liners",
           subtitle = "Half of it, that is...",
           width = "50%",
           header_color = "#51288590")
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/bbreg_samp.csv"
samp = read.csv(url)
samp = samp[,-1]
```

<br>

## Telling stories

- What just happened?

- Are these results a problem? 

  - Is the context behind them bad?
  
- What is data?

- What is a statistic?

  - What is it for?
  
  - What are the limitations of statistics? Of science?
  
<br>

- A major step in telling a story

  - Having an idea for the plot
  
- What makes a good plot?

<br>

- How do scientific inquiries happen?

  - By accident ([Grokking](https://arxiv.org/pdf/2201.02177))
  
  - General madness ([Game theory](https://www.cs.upc.edu/~ia/nash51.pdf))
  
  - Taking risks ([Chemotherapy and Emil Freireich](https://laskerfoundation.org/winners/combination-chemotherapy-for-lymphoma-and-acute-leukemia/#_))
  
  - Spite ([Scale-free networks](https://liorpachter.wordpress.com/2014/02/10/the-network-nonsense-of-albert-laszlo-barabasi/))
  
  - "Things breaking" (?)
  
<br>

- Is Fred Olsen a potent vector of norovirus?

  - Are Carnival cruises really that *safe* of an option?
  
<br>

- Bayesian methods

    - Information deficiency
    
    - "Automatic inference"
    
    - Assumptions $\approx$ data
    
$$
\begin{aligned}
& y \sim \text{Beta}(\alpha,\beta), \\
\\
&\alpha = \mu \phi, \\
\\
& \beta = (1- \mu) \phi, \\
\\
& g(\mu) = a + bx, \\
\\
& a,b \sim N(0,0.1), \\
\\
& \phi \sim \text{Gamma}(1,0.01).
\end{aligned}
$$

```{r realcruisechunk5, eval=FALSE, fig.align='center', message=FALSE, warning=FALSE}
bbreg = function(){
  
  # process model
  for(i in 1:n){
    y[i] ~ dbeta(alpha[i], beta[i])
    
    # mean/precision parameterization
    alpha[i] <- mu[i] * phi
    beta[i]  <- (1-mu[i]) * phi
    
    # logit link function
    logit(mu[i]) <- a + b*x[i]
  }
  
  # priors
  phi ~ dgamma(1,.01)
  a ~ dnorm(0,.1)
  b ~ dnorm(0,.1)
  
}

# setting up jags data
jags_dat = list(y = cruise$infected_proportion,
                x = as.integer(factor(cruise$liner)),
                n = nrow(cruise))

# initial proposal values
inits = list(a = .01, b = .01, phi = 1)

# fitting the model
bbreg_mcmc = jags.fit(jags_dat,c("a","b","phi"),
                      bbreg,inits,n.adapt=5000,
                      n.update=5000,n.iter=100000,
                      thin=1,n.chains=1)
```

```{r cruisechunk6, warning=FALSE,message=FALSE,fig.align='center'}
# reproducibility seed
set.seed(73)

# number of posterior samples
K = nrow(samp)

# set to carnival and fred olsen
x_pred = c(3,8)
a_post = samp[,1] # intercept post
b_post = samp[,2] # effect par post
phi_post = samp[,3] # precision par post

# vectorized posterior logit(mu) predictions
eta_post = outer(a_post, x_pred,
                 FUN = function(ai, x) ai) + 
  outer(b_post, x_pred)

# "un-do" the link function
mu_post = invlogit(eta_post)

# alpha posterior predictions
alpha_post = mu_post * phi_post

# beta posterior predictions
beta_post = (1 - mu_post) * phi_post

# posterior predictions of y
y_post = matrix(rbeta(length(alpha_post),
                      alpha_post, beta_post),
                nrow = K, ncol = length(x_pred))
```

```{r cruisechunk7,echo=FALSE,warning=FALSE,message=FALSE,fig.align='center'}
hist(y_post[,1],
     xlab = "Proportion infected",
     main = "Posterior prediction of Carnival outbreaks",
     col = "white")
hist(y_post[,2],
     xlab = "Proportion infected",
     main = "Posterior prediction of Fred Olsen outbreaks",
     col = "white")
```

```{r cruisechunk8, echo=FALSE,warning=FALSE,message=FALSE,fig.align='center'}
cred_int = t(apply(y_post, 2, quantile, probs = c(0.025, 0.5, 0.975)))

op = par(mar = c(5, 10, 4, 2))

plot(
  NULL,
  xlim = range(c(lo_r, hi_r)),
  ylim = c(0.5, length(levs) + 0.5),
  yaxt = "n",
  xlab = "Predicted infected proportion (95% CI)",
  ylab = "",
  main = "Predicted mean infection by liner"
)

segments(lo_r, seq_along(levs), hi_r, seq_along(levs), lwd = 3)
segments(cred_int[1,1], 3, cred_int[1,3], 3, lwd = 4, col = "green4")
segments(cred_int[2,1], 8, cred_int[2,3], 8, lwd = 4, col = "green4")
points(cred_int[1,2], 3, cex = 2, pch = "*")
points(cred_int[2,2], 8, cex = 2, pch = "*")

axis(2, at = seq_along(levs), labels = levs, las = 1, cex.axis = 0.75)

par(op)
```

- Supplemental material

  - Project proposal samples
  
  - STAT 341 dictionary
  
  - Varying R resources
  
- Extra reading

  - R for Data Science
  
  - Linear Models in R
  
- Assignment 2
  
<br>

## Go away