# Day 16

## Review

- Model Diagnostics

<br>

- Simulating data

```{r}
# simulating data for testing
set.seed(1)
n = 120
x = runif(n, 1, 5)
y = 3 + runif(1,1,3)*x + rnorm(n,0,2)
mtest = lm(y ~ x)
plot(x,y)
points(x, mtest$fitted.values, type = "l", lwd = 2, col = "red")
```

<br>

```{r echo=FALSE, message=FALSE, warning=FALSE}
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/framingham_heart_study.csv"
framingham = read.csv(url, stringsAsFactors = FALSE)
framingham = na.omit(framingham)
m1 = lm(currentSmoker ~ age + male + BMI + prevalentHyp, data = framingham)
```

<br>

- Linearity

```{r}
par(mfrow = c(1,2), mar = c(4,4,2,1))
plot(mtest, which = 1)
plot(m1, which = 1)
```

- Independence

    - What is the suspected dependence structure?
    
<br>

- Dependence on the response

<br>

- Dependence on a predictor

```{r}
par(mfrow = c(1,2), mar = c(4,4,2,1))
plot(x, mtest$residuals)
plot(framingham$age, m1$residuals)
```

<br>

- Another useful quick check

<br>


- Homoskedasticity

```{r}
par(mfrow = c(1,2), mar = c(4,4,2,1))
plot(mtest, which = 3)
plot(m1, which = 3)
```

<br>

- Normality

```{r}
par(mfrow = c(1,2), mar = c(4,4,2,1))
plot(mtest, which = 2)
plot(m1, which = 2)
```

<br>

- Outliers

<br>

```{r}
# generating new data with some extreme outliers
ynew = c(y,300,250,188,500)
xnew = c(x,0,1,3,2)
mtest2 = lm(ynew ~ xnew)

par(mfrow = c(1,2), mar = c(4,4,2,1))
plot(mtest, which = 5)
plot(mtest2, which = 5)
```

```{r}
plot(m1, which = 5)
```

<br>

## Nonlinear models

```{r}
plot(m1, which = 1)
```

- Logistic regression

$$
p(y) = \frac{1}{1 + \exp\{-(\beta_0 + \beta_1 x\}}
$$

- Fitting the model requires a different algorithm

    - "Convergence" (and [failure to converge](https://en.wikipedia.org/wiki/Rastrigin_function))

```{r}
m2 = nls(currentSmoker ~ 1 / 
           (1 + exp(-(b0 + b1 * age +
                        b2 * male +
                        b3 * BMI +
                        b4 * prevalentHyp))),
         data  = framingham,
         start = list(b0 = 0, b1 = 0, b2 = 0, b3 = 0, b4 = 0))
summary(m2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

means = colMeans(na.omit(framingham[, c("age", "male", "BMI", "prevalentHyp")]))

age_seq = seq(min(framingham$age, na.rm = TRUE), max(framingham$age, na.rm = TRUE), length.out = 200)
age_new = data.frame(age = age_seq, male = means["male"],
                      BMI = means["BMI"], prevalentHyp = means["prevalentHyp"])
age_pred = predict(m2, newdata = age_new)

plot(framingham$age, framingham$currentSmoker,
     pch = 16,
     xlab = "Age", ylab = "P(Smoker)",
     main = "", ylim = c(0, 1))
lines(age_seq, age_pred, col = "firebrick", lwd = 2)

bmi_seq = seq(min(framingham$BMI, na.rm = TRUE), max(framingham$BMI, na.rm = TRUE), length.out = 200)
bmi_new = data.frame(age = means["age"], male = means["male"],
                      BMI = bmi_seq, prevalentHyp = means["prevalentHyp"])
bmi_pred = predict(m2, newdata = bmi_new)

plot(framingham$BMI, framingham$currentSmoker,
     pch = 16,
     xlab = "BMI", ylab = "P(Smoker)",
     main = "", ylim = c(0, 1))
lines(bmi_seq, bmi_pred, col = "firebrick", lwd = 2)

par(mfrow = c(1, 1))
```

<br>

- Additive models

$$
y_i = f(x_i) + \epsilon_i
$$

$$
\epsilon_i \sim N(0, \sigma^2)
$$

<br>

```{r message=FALSE, warning=FALSE}
library(mgcv)
m3 = gam(currentSmoker ~ s(age) + male + s(BMI) + prevalentHyp, data = framingham)
summary(m3)
```

<br>

```{r echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
means = colMeans(na.omit(framingham[, c("age", "male", "BMI", "prevalentHyp")]))

age_seq = seq(min(framingham$age, na.rm = TRUE), max(framingham$age, na.rm = TRUE), length.out = 200)
age_new = data.frame(age = age_seq, male = means["male"],
                     BMI = means["BMI"], prevalentHyp = means["prevalentHyp"])
age_pred = predict(m3, newdata = age_new)
plot(framingham$age, framingham$currentSmoker,
     pch = 16,
     xlab = "Age", ylab = "P(Smoker)",
     main = "", ylim = c(0, 1))
lines(age_seq, age_pred, col = "firebrick", lwd = 2)

bmi_seq = seq(min(framingham$BMI, na.rm = TRUE), max(framingham$BMI, na.rm = TRUE), length.out = 200)
bmi_new = data.frame(age = means["age"], male = means["male"],
                     BMI = bmi_seq, prevalentHyp = means["prevalentHyp"])
bmi_pred = predict(m3, newdata = bmi_new)
plot(framingham$BMI, framingham$currentSmoker,
     pch = 16,
     xlab = "BMI", ylab = "P(Smoker)",
     main = "", ylim = c(0, 1))
lines(bmi_seq, bmi_pred, col = "firebrick", lwd = 2)

par(mfrow = c(1, 1))
```

<br>

- How do we compare these models?

<br>

- Coefficient of Determination

$$
R^2 = \frac{\sum_{i=1}^n (\hat{y} - \bar{y})^2}{\sum_{i=1}^n (y - \bar{y})^2}
$$

<br>

$$
R^2_{adj} = 1- (1 - R^2)\frac{n-1}{n-p-1}
$$

<br>

```{r}
summary(m1)$r.squared
# logistic regression needs a psuedo r squared (pain)
summary(m3)$r.sq
```

<br>

<br>

**Any questions?**

<br>

<br>

## Correlated Errors

- Dependence structures

    - Dependent on predictions
    
    - Dependent on a predictor

- The most common predictors to see dependenc in?

    - Spatial location
    
    - Time
    
    
```{r}
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/TN05.csv"
TN15 = read.csv(url)
TN15a = TN15 %>%
  filter(!is.na(Long), !is.na(Total_N)) %>%
  group_by(SITE_NAME) %>%
  summarise(Total_N = sum(Total_N),
            across(everything(), first),
            .groups = "drop")


par(mfrow = c(2,1), mar = c(4.5,4.5,1,1))
plot(TN15a$Lat, TN15a$Total_N, xlab = "Latitude", ylab = "Total Nitrogen")
plot(TN15a$Long, TN15a$Total_N, xlab = "Longitude", ylab = "Total Nitrogen")
```

<br>

- Assuming one dimension

<br>

- Intercept-only linear model

```{r}
m1 = lm(Total_N ~ 1, data = TN15a)
plot(sort(TN15a$Long), residuals(m1), col = "#D1D1D1")
abline(h = 0)
```

<br>

- Intercept only spatial linear model

```{r}
library(nlme)
m2 = gls(Total_N ~ 1, data = TN15a, correlation = corGaus(form = ~ Long, nugget = TRUE, value = c(1, 0.05), fixed = FALSE), method = "ML")
summary(m2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
phi_TN15    = exp(m2$modelStruct$corStruct[1])^2
nugget_TN15 = 1 / (1 + exp(-m2$modelStruct$corStruct[2]))
X_TN15 = rep(1, length(TN15a$Total_N))
y_TN15 = TN15a$Total_N
 
D_TN15 = as.matrix(dist(TN15a$Long, diag = TRUE, upper = TRUE))
 
sigma2.eta_TN15 = (m2$sigma^2) * (1 - nugget_TN15)
sigma2.e_TN15   = (m2$sigma^2) * nugget_TN15
 
R_TN15 = exp(-D_TN15^2 / phi_TN15)
I_TN15 = diag(1, length(y_TN15))
V_TN15 = sigma2.eta_TN15 * R_TN15 + sigma2.e_TN15 * I_TN15
beta_TN15 = solve(t(X_TN15) %*% solve(V_TN15) %*% X_TN15) %*%
             t(X_TN15) %*% solve(V_TN15) %*% y_TN15
 
eta_TN15  = sigma2.eta_TN15 * R_TN15 %*% solve(V_TN15) %*%
             (y_TN15 - X_TN15 %*% beta_TN15)
 
fitted_TN15 = X_TN15 %*% beta_TN15 + eta_TN15
e.hat_TN15  = y_TN15 - fitted_TN15

par(mar = c(5.1, 4.1, 2.1, 2.1), oma = c(0, 0, 0, 0))
par(mfrow = c(1, 2))
par(pch = 19)
 
plot(TN15a$Long, TN15a$Total_N,
     xlab = "Longitude", ylab = "Total Nitrogen (mg/L)",
     las = 1, col = rgb(0, 0, 0, 0.25))
lines(TN15a$Long[order(TN15a$Long)], fitted_TN15[order(TN15a$Long)], col = "black")
abline(h = beta_TN15, col = "purple", lty = 2)
 
plot(TN15a$Long, e.hat_TN15,
     xlab = "Longitude", ylab = expression(hat(epsilon)),
     las = 1, col = rgb(0, 0, 0, 0.25))
abline(h = 0, col = "grey50", lty = 2)
```

<br>

- 2D Models

<br>

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(sf)
library(maps)
library(dplyr)

kansas_map = map_data("state") |>
  filter(region == "kansas")

kansas_sf = st_as_sf(
  kansas_map,
  coords = c("long", "lat"),
  crs    = 4326
)

points_sf = st_as_sf(
  TN15a,
  coords = c("Long", "Lat"),
  crs    = 4326
)

ggplot() +
    geom_polygon(
    data    = kansas_map,
    aes(x = long, y = lat, group = group),
    fill    = "white",
    colour  = "#9e9e9e",
    linewidth = 0.5
  ) +
  
  geom_point(
    data  = TN15a,
    aes(x = Long, y = Lat,
        colour = Total_N),
    alpha = 0.85
  ) +
  
  scale_colour_gradient(
    low    = "#D3D3FF",
    high   = "#512885",
    name   = "Total Nitrogen"
  ) +
  
  scale_size_continuous(
    range = c(3, 9),
    name  = "Total Nitrogen"
  ) +
  
  coord_fixed(
    ratio  = 1.3,
    xlim   = c(-102.5, -94.5),
    ylim   = c(36.8,   40.2)
  ) +
  
  labs(
    title    = "",
    subtitle = "",
    caption  = "",
    x        = "Longitude",
    y        = "Latitude"
  ) +
  
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(colour = "grey40", size = 11),
    legend.position = "right",
    panel.grid      = element_line(colour = "grey90", linewidth = 0.3)
  )
```

<br>

- Kriging

    - Result of a South African statistician working for a mining corporation
    
    - "How many bore holes do we need before we're certain where the gold is?"
    
- Weighting regression by distance

- Assumes space arises from a Gaussian Process

- Computationally expensive (cubic time)

    - We just approximate it
    
<br>

- GAMs make a return!

```{r}
library(mgcv)
m3 = gam(Total_N ~ te(Lat, Long, bs = "gp"), data = TN15a)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
pred_grid = expand.grid(
  Long = seq(-102.5, -94.5, length.out = 200),
  Lat  = seq(36.8,   40.2,  length.out = 200)
)

pred_grid$Total_N = predict(m3, newdata = pred_grid)

kansas_map = map_data("state") |> filter(region == "kansas")

kansas_poly = st_polygon(
  list(cbind(kansas_map$long, kansas_map$lat))
) |>
  st_sfc(crs = 4326)

grid_sf = st_as_sf(pred_grid, coords = c("Long", "Lat"), crs = 4326)
inside  = st_within(grid_sf, kansas_poly, sparse = FALSE)[ , 1]

pred_clipped = pred_grid[inside, ]

ggplot() +

  geom_tile(
    data = pred_clipped,
    aes(x = Long, y = Lat, fill = Total_N)
  ) +

  geom_polygon(
    data      = kansas_map,
    aes(x = long, y = lat, group = group),
    fill      = NA,
    colour    = "grey30",
    linewidth = 0.5
  ) +

  scale_fill_gradient(
    low  = "#D3D3FF",
    high = "#512885",
    name = "Total Nitrogen"
  ) +

  coord_fixed(
    ratio = 1.3,
    xlim  = c(-102.5, -94.5),
    ylim  = c(36.8,   40.2)
  ) +

  labs(
    title = "Interpolated Total Nitrogen",
    x     = "Longitude",
    y     = "Latitude"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    legend.position = "right",
    panel.grid      = element_line(colour = "grey90", linewidth = 0.3)
  )
```

<br>

<br>

## Information Criterion

- AIC, BIC, CIC, DIC, ..., ZIC

- These are different from $R^2$ 

    - Diet coke is different from coke
    
- These are methods of assessing models under *certain shared assumptions*

<br>

**Akaike's Information Criterion (AIC)**

- Log-likelihood scoring rule

    - You cannot log transform your data and compare it using AIC the same way as untransformed
    
- AIC is best at scoring similar families and structures of models

- Variable rejection/selection? Good.

- Difference between binomial regression with an inverse logit link vs poisson with a log link? 

<br>

```{r}
# "lower is better"
AIC(m1)
AIC(m2)
AIC(m3)
```

<br>

## Go away



