# Day 17

## Review

- Spatial models

```{r}
library(nlme)
m2 = gls(Total_N ~ 1, data = TN15a, correlation = corGaus(form = ~ Long, nugget = TRUE, value = c(1, 0.05), fixed = FALSE), method = "ML")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
phi_TN15    = exp(m2$modelStruct$corStruct[1])^2
nugget_TN15 = 1 / (1 + exp(-m2$modelStruct$corStruct[2]))
X_TN15 = rep(1, length(TN15a$Total_N))
y_TN15 = TN15a$Total_N
 
D_TN15 = as.matrix(dist(TN15a$Long, diag = TRUE, upper = TRUE))
 
sigma2.eta_TN15 = (m2$sigma^2) * (1 - nugget_TN15)
sigma2.e_TN15   = (m2$sigma^2) * nugget_TN15
 
R_TN15 = exp(-D_TN15^2 / phi_TN15)
I_TN15 = diag(1, length(y_TN15))
V_TN15 = sigma2.eta_TN15 * R_TN15 + sigma2.e_TN15 * I_TN15
beta_TN15 = solve(t(X_TN15) %*% solve(V_TN15) %*% X_TN15) %*%
             t(X_TN15) %*% solve(V_TN15) %*% y_TN15
 
eta_TN15  = sigma2.eta_TN15 * R_TN15 %*% solve(V_TN15) %*%
             (y_TN15 - X_TN15 %*% beta_TN15)
 
fitted_TN15 = X_TN15 %*% beta_TN15 + eta_TN15
e.hat_TN15  = y_TN15 - fitted_TN15

par(mar = c(5.1, 4.1, 2.1, 2.1), oma = c(0, 0, 0, 0))
par(mfrow = c(1, 2))
par(pch = 19)
 
plot(TN15a$Long, TN15a$Total_N,
     xlab = "Longitude", ylab = "Total Nitrogen (mg/L)",
     las = 1, col = rgb(0, 0, 0, 0.25))
lines(TN15a$Long[order(TN15a$Long)], fitted_TN15[order(TN15a$Long)], col = "black")
abline(h = beta_TN15, col = "purple", lty = 2)
 
plot(TN15a$Long, e.hat_TN15,
     xlab = "Longitude", ylab = expression(hat(epsilon)),
     las = 1, col = rgb(0, 0, 0, 0.25))
abline(h = 0, col = "grey50", lty = 2)
```

- 2D models

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(sf)
library(maps)
library(dplyr)

kansas_map = map_data("state") |>
  filter(region == "kansas")

kansas_sf = st_as_sf(
  kansas_map,
  coords = c("long", "lat"),
  crs    = 4326
)

points_sf = st_as_sf(
  TN15a,
  coords = c("Long", "Lat"),
  crs    = 4326
)

ggplot() +
    geom_polygon(
    data    = kansas_map,
    aes(x = long, y = lat, group = group),
    fill    = "white",
    colour  = "#9e9e9e",
    linewidth = 0.5
  ) +
  
  geom_point(
    data  = TN15a,
    aes(x = Long, y = Lat,
        colour = Total_N),
    alpha = 0.85
  ) +
  
  scale_colour_gradient(
    low    = "#D3D3FF",
    high   = "#512885",
    name   = "Total Nitrogen"
  ) +
  
  scale_size_continuous(
    range = c(3, 9),
    name  = "Total Nitrogen"
  ) +
  
  coord_fixed(
    ratio  = 1.3,
    xlim   = c(-102.5, -94.5),
    ylim   = c(36.8,   40.2)
  ) +
  
  labs(
    title    = "",
    subtitle = "",
    caption  = "",
    x        = "Longitude",
    y        = "Latitude"
  ) +
  
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(colour = "grey40", size = 11),
    legend.position = "right",
    panel.grid      = element_line(colour = "grey90", linewidth = 0.3)
  )
```

- Kriging
    
    - Weighting regression by distance

    - Assumes space arises from a Gaussian Process

    - Computationally expensive (cubic time)

```{r}
library(mgcv)
m3 = gam(Total_N ~ te(Lat, Long, bs = "gp"), data = TN15a)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
pred_grid = expand.grid(
  Long = seq(-102.5, -94.5, length.out = 200),
  Lat  = seq(36.8,   40.2,  length.out = 200)
)

pred_grid$Total_N = predict(m3, newdata = pred_grid)

kansas_map = map_data("state") |> filter(region == "kansas")

kansas_poly = st_polygon(
  list(cbind(kansas_map$long, kansas_map$lat))
) |>
  st_sfc(crs = 4326)

grid_sf = st_as_sf(pred_grid, coords = c("Long", "Lat"), crs = 4326)
inside  = st_within(grid_sf, kansas_poly, sparse = FALSE)[ , 1]

pred_clipped = pred_grid[inside, ]

ggplot() +

  geom_tile(
    data = pred_clipped,
    aes(x = Long, y = Lat, fill = Total_N)
  ) +

  geom_polygon(
    data      = kansas_map,
    aes(x = long, y = lat, group = group),
    fill      = NA,
    colour    = "grey30",
    linewidth = 0.5
  ) +

  scale_fill_gradient(
    low  = "#D3D3FF",
    high = "#512885",
    name = "Total Nitrogen"
  ) +

  coord_fixed(
    ratio = 1.3,
    xlim  = c(-102.5, -94.5),
    ylim  = c(36.8,   40.2)
  ) +

  labs(
    title = "Interpolated Total Nitrogen",
    x     = "Longitude",
    y     = "Latitude"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    legend.position = "right",
    panel.grid      = element_line(colour = "grey90", linewidth = 0.3)
  )
```

<br>

- Food for thought

    - Geographic coordinates take up 2 dimensions (x and y)
    
    - Height / elevation typically takes the third (z)
    
    - What about a 2D object across multiple time steps?
    
<br>

## Temporal Statistics

```{r}
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/GlobalLandTemperaturesByState.csv"
land_temp_state = na.omit(read.csv(url, stringsAsFactors = FALSE))
ks_temps = subset(land_temp_state, land_temp_state$State == "Kansas")
ks_temps = ks_temps[order(ks_temps$dt), ]
head(ks_temps)
```

```{r message=FALSE, warning=FALSE}
library(lubridate)
ks_temps$dt = as.Date(ks_temps$dt)
ks_temps$difft = as.numeric(difftime(ks_temps$dt, min(ks_temps$dt), units = "days"))
ks_temps$month = as.integer(format(ks_temps$dt, "%m"))
plot(ks_temps$difft, ks_temps$AverageTemperature, pch = 16, col  = "#D1D1D1",
     xlab = "Days since 09-01-1768", ylab = "Avg Temp (°C)")
```

```{r}
y = ks_temps$AverageTemperature
x = ks_temps$difft
lm(y ~ x)
```

- Moving (rolling) average

$$
\frac{1}{k} \sum_{i = n - k + 1}^n p_i
$$

```{r echo=FALSE, message=FALSE, warning=FALSE}
rolling_avg = function(x, k) {
  n = length(x)
  half_k = floor(k / 2)
  out = rep(NA, n)
  for (i in (half_k + 1):(n - half_k)) {
    out[i] = mean(x[(i - half_k):(i + half_k)])
  }
  out
}

window_size = 12
ks_roll_avg = rolling_avg(y, k = window_size)

plot(x, y,
     col  = "#D1D1D1", pch = 16,
     xlab = "Days since 09-01-1768",
     ylab = "Avg Temp (°C)")
lines(x, ks_roll_avg, col = "red", lwd = 2)
```

- Footnote on Autoregression / white noise / stationarity
    
<br>

## Collinearity

```{r message=FALSE, warning=FALSE}
library(httr)
library(readxl)

url = "https://doi.org/10.1371/journal.pone.0148743.s005"    
GET(url, write_disk(path <- tempfile(fileext = ".xls")))
```

```{r message=FALSE, warning=FALSE}
df1 = read_excel(path=path,sheet = "Data", col_names=TRUE,col_types="numeric")
notes = read_excel(path=path,sheet = "Notes", col_names=FALSE)
```

```{r}
head(df1)
```

```{r}
notes
```

```{r}
m1 = lm(CDEP ~ NCAT + WPOP,data=df1)
summary(m1)
```

```{r}
m2 = lm(CDEP ~ NCAT + WPOP + WKIL,data=df1)
summary(m2)
```

```{r}
cor(df1[,c(4,5,7)],use="complete.obs")
```

<br>

## More about variance

- Centered data

```{r}
head(faithful)
y = faithful$eruptions
x = faithful$waiting
summary(lm(y ~ x)) # model with no centering

xc = scale(x, center = TRUE, scale = FALSE)
mean(xc) # zero mean 
summary(lm(y ~ xc)) # centered model
```

- What's going on here?

```{r}
Xc = cbind(1,xc)
solve(t(Xc)%*%Xc)%*%t(Xc)%*%y
mean(y)
```

<br>

## Constant variance

```{r}
library(gamair)
data(hubble)

m1 = lm(y ~ x - 1, data = hubble)
confint(m1)
```

```{r}
# Plot data and E(y)
plot(hubble$x,hubble$y,xlab="Distance (Mpc)",
ylab=expression("Velocity (km"*s^{-1}*")"),pch=20)
abline(m1)
```

```{r}
# Plot residuals vs. x
e.hat = residuals(m1)
hist(e.hat,col="grey",breaks=15,main="",xlab=expression(hat(epsilon)))  
plot(hubble$x,e.hat,xlab="Distance(Mpc)",ylab=expression(hat(epsilon)),pch=20)
```

```{r}
# Plot prediction intervals
y.pi = predict(m1,newdata=data.frame(x = 0:25),interval = "predict")
plot(hubble$x,hubble$y,xlab="Distance (Mpc)",
ylab=expression("Velocity (km"*s^{-1}*")"),xlim=c(0,25),ylim=c(0,2500),pch=20)
points(0:25,y.pi[,2],typ="l",lwd=2,col="purple")
points(0:25,y.pi[,3],typ="l",lwd=2,col="purple")
```

<br>

```{r}
library(nlme)
m2 = gls(y~x-1,weights = varExp(form=~x),data=hubble)
summary(m2)
confint(m2)
```

```{r}
# Plot prediction intervals
X = model.matrix(y~x-1,data=hubble)
sigma2.hat = summary(m1)$sigma^2
sigma2.hat*solve(t(X)%*%X)
vcov(m1)
```

```{r echo=FALSE}
X = model.matrix(y~x-1,data=hubble)
beta.hat = coef(m2)
sigma2.hat = summary(m2)$sigma^2
theta.hat = summary(m2)$modelStruct$varStruct
V = diag(c(sigma2.hat*exp(2*X%*%theta.hat)))
X.new = model.matrix(~x-1,data=data.frame(x = 1:25))
y.pi = cbind(X.new%*%beta.hat - qt(0.975,24-1)*sqrt(X.new^2%*%solve(t(X)%*%solve(V)%*%X) + sigma2.hat*exp(2*X.new%*%theta.hat)),
X.new%*%beta.hat + qt(0.975,24-1)*sqrt(X.new^2%*%solve(t(X)%*%solve(V)%*%X) + sigma2.hat*exp(2*X.new%*%theta.hat)))

plot(hubble$x,hubble$y,xlab="Distance (Mpc)",
 ylab=expression("Velocity (km"*s^{-1}*")"),xlim=c(0,25),ylim=c(0,2500),pch=20)
points(1:25,y.pi[,1],typ="l",lwd=2,col="purple")
points(1:25,y.pi[,2],typ="l",lwd=2,col="purple")
```

<br>

## Prediction Interval Coverage

```{r message=FALSE, warning=FALSE}
library(mgcv)
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/SOL_PHOS.csv"
SOL_PHOS = read.csv(url)[,-1]

SOL_PHOSTET = SOL_PHOS[SOL_PHOS$year %in% seq(1973, 2024, 2), ] # odd yrs
SOL_PHOSTRT = SOL_PHOS[SOL_PHOS$year %in% seq(1974, 2024, 2), ] # even yrs

lat_mid = median(SOL_PHOS$Lat_j)

SOL_PHOSTEH = SOL_PHOS[SOL_PHOS$Lat_j >= lat_mid, ]
SOL_PHOSTRH = SOL_PHOS[SOL_PHOS$Lat_j <  lat_mid, ]


coverage_mlr = function(model, test_data, alpha = 0.05) {
  preds = predict(model, newdata = test_data, interval = "prediction", 
                   level = 1 - alpha)
  covered = test_data$PHOSPHU >= preds[, "lwr"] & 
    test_data$PHOSPHU <= preds[, "upr"]
  mean(covered, na.rm = TRUE)
}
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
coverage_gam = function(model, test_data, alpha = 0.05, n_sim = 1000) {
  preds  = predict(model, newdata = test_data, type = "link", se.fit = TRUE)
  
  
  set.seed(42)
  sims = matrix(
    rnorm(nrow(test_data) * n_sim, mean = preds$fit, sd = preds$se.fit),
    nrow = nrow(test_data), ncol = n_sim
  )
  
  
  sims_resp = model$family$linkinv(sims)
  
  
  phi = summary(model)$dispersion
  p   = model$family$getTheta(trans = TRUE)
  
  
  mu_mat = sims_resp
  shape  = mu_mat^(2 - p) / phi
  scale  = phi * mu_mat^(p - 1)
  obs_noise = matrix(
    rgamma(length(mu_mat), shape = pmax(shape, 0.001), 
           scale = pmax(scale, 0.001)),
    nrow = nrow(test_data)
  )
  
  lwr = apply(obs_noise, 1, quantile, probs = alpha / 2,     na.rm = TRUE)
  upr = apply(obs_noise, 1, quantile, probs = 1 - alpha / 2, na.rm = TRUE)
  
  covered = test_data$PHOSPHU >= lwr & test_data$PHOSPHU <= upr
  mean(covered, na.rm = TRUE)
}
```

```{r}
# spatio-temporal multiple linear regression 
mlrt_SUB = lm(PHOSPHU ~ doy + year + Lat_j + Long_j, data = SOL_PHOSTRT)
mlrh_SUB = lm(PHOSPHU ~ doy + year + Lat_j + Long_j, data = SOL_PHOSTRH)

# spatio-temporal tensors fit with tweedie gams
tpt_SUB = gam(PHOSPHU ~ te(Lat_j, Long_j, doy,
                            bs = c("tp", "tp", "cc")) + year,
               family = tw(), data = SOL_PHOSTRT, method = "REML")
tph_SUB = gam(PHOSPHU ~ te(Lat_j, Long_j, doy,
                            bs = c("tp", "tp", "cc")) + year,
               family = tw(), data = SOL_PHOSTRH, method = "REML")
```

```{r}
coverage_mlr(mlrt_SUB, SOL_PHOSTRT, alpha = 0.1)
coverage_gam(tpt_SUB, SOL_PHOSTRT, alpha = 0.1)

coverage_mlr(mlrh_SUB, SOL_PHOSTRH, alpha = 0.1)
coverage_gam(tph_SUB, SOL_PHOSTRH, alpha = 0.1)
```

<br>

<br>

## Go away



