# Day 18

## Announcements

- Plans for next week

    - Finish distributional assumptions

    - Course review (you made it!)

- Datafest

- Grades

<br>

## Review

Whiteboard

- Assumptions of the linear model

<br>

- Assumptions being "violated"

<br>

- "Fixing" the broken assumptions

<br>

- Methods of assessing / checking models

<br>

- Temporal statistics

```{r}
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/GlobalLandTemperaturesByState.csv"
land_temp_state = na.omit(read.csv(url, stringsAsFactors = FALSE))
ks_temps = subset(land_temp_state, land_temp_state$State == "Kansas")
ks_temps = ks_temps[order(ks_temps$dt), ]
head(ks_temps)
```

```{r message=FALSE, warning=FALSE}
library(lubridate)
ks_temps$dt = as.Date(ks_temps$dt)
ks_temps$difft = as.numeric(difftime(ks_temps$dt, min(ks_temps$dt), units = "days"))
ks_temps$month = as.integer(format(ks_temps$dt, "%m"))
plot(ks_temps$difft, ks_temps$AverageTemperature, pch = 16, col  = "#D1D1D1",
     xlab = "Days since 09-01-1768", ylab = "Avg Temp (°C)")
```

```{r}
y = ks_temps$AverageTemperature
x = ks_temps$difft
lm(y ~ x)
```

- Moving (rolling) average

$$
\frac{1}{k} \sum_{i = n - k + 1}^n p_i
$$

```{r echo=FALSE, message=FALSE, warning=FALSE}
rolling_avg = function(x, k) {
  n = length(x)
  half_k = floor(k / 2)
  out = rep(NA, n)
  for (i in (half_k + 1):(n - half_k)) {
    out[i] = mean(x[(i - half_k):(i + half_k)])
  }
  out
}

window_size = 12
ks_roll_avg = rolling_avg(y, k = window_size)

plot(x, y,
     col  = "#D1D1D1", pch = 16,
     xlab = "Days since 09-01-1768",
     ylab = "Avg Temp (°C)")
lines(x, ks_roll_avg, col = "red", lwd = 2)
```

<br>

**Any questions?**

<br>

## Spatio-temporal data

- Simpson's paradox

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)

set.seed(42)
n <- 1000
group_a <- data.frame(
  x = rnorm(n, mean = 2, sd = 0.5),
  y = rnorm(n, mean = 4, sd = 0.5),
  group = "A"
)
group_b <- data.frame(
  x = rnorm(n, mean = 4, sd = 0.5),
  y = rnorm(n, mean = 2, sd = 0.5),
  group = "B"
)
group_a$y <- group_a$y + 0.6 * (group_a$x - mean(group_a$x))
group_b$y <- group_b$y + 0.6 * (group_b$x - mean(group_b$x))
data <- rbind(group_a, group_b, )

par(mar = c(4,4,1,1))
plot(data$x, data$y, xlab = "X", ylab = "Y", pch = 16, col = "#00000080")

```

<br>

- Change of support

<br>

- Kriging methods

    - Ordinary 

    - Universal 

    - Spatio-temporal 

<br>

- [Krige-and-regress](https://onlinelibrary.wiley.com/doi/10.1002/env.888)

<br>

- Reality of spatio-temporal data

    - Spatial statistics isn't very useful
    
    - Temporal statistics is highly assumption driven
    
    - Spatio-temporal data is everywhere
    
    - Everything hits the fan so quickly

- Suggestion: split the problem to piece you understand

    - Make time a non-issue (divide and conquer)
    
    - Make space a non-issue (clustering)
    
- Spatial statistics tends to be more intuitive
    
    - Adding time to space isn't as hard as the converse
    
- STAT 764, Spring 2028 

    - If you kept up decently well before
    
<br>

## EDA

- Live example

<br>

## In-class work time

<br>

<br>

<br>

<br>

## Go away
