# Day 20

## Review

- Binomial distribution

  - Discrete, binary processes
  
  - Success/failure

\[
X \sim \text{Bin}(n,p)
\qquad \qquad 
\mathbb{E}X = np
\qquad \qquad 
\mathbb{V}X = np(1-p)
\]

- Poisson distribution

  - Positive discrete counts
  
\[
X \sim \text{Pois}(\lambda)
\qquad \qquad 
\mathbb{E}X = \mathbb{V}X = \lambda
\]

- Geometric distribution

  - Number of trials to reach success
  
\[
X \sim \text{Geom}(p) 
\qquad \qquad
\mathbb{E}X = 1/p
\]

- Hypergeometric distribution

  - Geometric distribution
  
  - Sampling without replacement
  
\[
X \sim \text{Hyper}(N,K,n) 
\qquad \qquad
\mathbb{E}X = n \frac{K}{N}
\]

<br>

- Continuous expected values

\[
\mathbb{E}X = \int_{\chi}x f_x(x)
\]

- Gamma distribution

  - Positive continuous values

\[
X \sim \text{Gamma}(\alpha,\theta)
\qquad \qquad 
\mathbb{E}X = \alpha \theta
\qquad \qquad 
\mathbb{V}X = \alpha \theta^2
\]

- Beta distribution

  - Proportions/probabilities

\[
X \sim \text{Beta}(\alpha,\beta)
\qquad \qquad 
\mathbb{E}X = \frac{\alpha}{\alpha + \beta}
\qquad \qquad 
\mathbb{V}X = \frac{\alpha \beta}{(\alpha + \beta)^2 (\alpha + \beta + 1)}
\]

- Uniform distribution

  - Equi-probable events
  
  - Defined boundaries (a and b)
  
\[
X \sim \text{Unif}(a,b)
\qquad \qquad 
\mathbb{E}X = \frac{1}{2}(b-a)
\qquad \qquad 
\mathbb{V}X = \frac{1}{12}(b-a)^2
\]


## A clarification

```{r message=FALSE, warning=FALSE}
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/heart.csv"
heart = read.csv(url, stringsAsFactors = FALSE)
```

```{r}
hist(heart$chol, breaks = 20, xlab = "Serum Cholesterol", main = "Heart Survey Participants", xlim = c(120,360))
```

- Define cholesterol as $y_i$

- *We assume* from the histogram

$$
y_i \sim N(\mu, \sigma^2)
$$

- We *added* the assumption there may be multiple normal distributions

$$
y_i \sim N(\mu_i, \sigma^2 \boldsymbol{I})
$$

- When we fit a regression model:

$$
\mu_i = \beta_0 + \beta_1x_{1_i} + \dots + \beta_px_{p_i}
$$

- As a reminder:

$$
\sigma^2 = \frac{1}{N}\sum_{i=1}^N (x - \mu)^2
$$

<br>

## Non-normal processes

### Binomial

- What is the process that creates a smoker?

<br>

```{r}
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/framingham_heart_study.csv"
framingham = na.omit(read.csv(url, stringsAsFactors = FALSE))
```

```{r}
table(framingham$currentSmoker)
```

<br>

$$
y \sim \text{Bin}(N, p)
$$

```{r}
set.seed(1)
rbinom(1,1,0.5)
rbinom(10,1,0.5)
```

```{r}
hist(rbinom(1000,5,0.3))
```

<br>

- We don't really care about $N$

    - $p$ is far more interesting
    
```{r}
m1 = lm(currentSmoker ~ age + BMI, data = framingham)
m2 = glm(currentSmoker ~ age + BMI, family = binomial(link = "logit"), data = framingham)
```

```{r}
summary(m1)
```

```{r}
summary(m2)
```

```{r}
df_pred = data.frame(age = seq(min(framingham$age), max(framingham$age), length.out = 1000),
                     BMI = seq(min(framingham$BMI), max(framingham$BMI), length.out = 1000))
df_pred$Ey = predict(m2, type = "response", newdata = df_pred)

plot(df_pred$age, df_pred$Ey, type = "l", col = "red", lwd = 1.5,
     xlab = "Age", ylab = "P(Smoker)")
points(df_pred$age, predict(m1, type = "response", newdata = df_pred), type = "l", lwd = 1.5)
```

<br>

## Link functions

$$
\begin{array}{lll}
\hline
\text{Distribution} & \text{Link Name} & \text{Formula} \\
\hline
\text{Normal}   & \text{Identity} & \mu \\
\text{Poisson}  & \text{Log}      & \ln(\mu) \\
\text{Binomial} & \text{Logit}    & \ln\left(\frac{\mu}{1-\mu}\right) \\
\text{Gamma}    & \text{Inverse}  & 1/\mu \\
\hline
\end{array}
$$

- Numerical stability

    - We love these in computational statistics*
    
    - *When they retain numerical stability

- Analytical convenience

    - For mathemtical statistics these functions cause problems
    
    - They solve as many as they create, but in the long run we try to escape them
    
<br>

### Poisson

- What determines how many you smoke per day?

```{r}
hist(framingham$cigsPerDay, xlab = "Cigarettes", main = "Cigarettes per day")
```

- Probably that you smoke...

```{r}
smokers = framingham[framingham$currentSmoker == 1,]
hist(smokers$cigsPerDay, xlab = "Cigarettes", main = "Cigarettes per day")
```

<br>

$$
y \sim Pois(\lambda)
$$

```{r}
hist(rpois(1000,5))
```

$$
\lambda = e^{\beta_0 + \beta_1x}
$$

<br>

```{r}
m1 = glm(cigsPerDay ~ smokers$heartRate, family = poisson(link = "log"), data = smokers)
summary(m1)
```

<br>

### Beta

```{r}
library(mgcv)
url = "https://raw.githubusercontent.com/rmshksu/teaching-data/refs/heads/main/cruisedata.csv"
cruise = read.csv(url, stringsAsFactors = FALSE)


par(mar = c(5, 9, 4, 2))
boxplot(cruise$infected_proportion ~ cruise$liner,
        xlab = "Proportion infected",
        ylab = "",
        main = "Norovirus outbreaks on cruise liners",
        horizontal = TRUE,
        las = 1,
        cex.axis = 0.7,
        col = "#D1D1D190")
```

\[
\left[y|\alpha, \beta \right] \equiv \text{Beta}(\alpha,\beta)
\]

\[
\alpha = \mu \phi \qquad \beta = (1 - \mu) \phi \qquad g(\mu) = \textbf{X}^\prime \boldsymbol{\beta}
\]

```{r}
m1 = gam(infected_proportion ~ as.factor(liner) - 1,
         data = cruise, family = betar(link = "logit"))
```

```{r, echo=FALSE,warning=FALSE,message=FALSE,fig.align='center'}
# liner levels exactly as used in the fitted model
levs <- levels(model.frame(m1)[["as.factor(liner)"]])

# prediction data: one row per liner
nd <- data.frame(liner = factor(levs, levels = levs))

# get fitted linear predictor + SE per liner
pr <- predict(m1, newdata = nd, type = "link", se.fit = TRUE)

# 95% CI on link scale, then transform to response
invlogit <- m1$family$linkinv
fit_r <- invlogit(pr$fit)
lo_r  <- invlogit(pr$fit - 1.96 * pr$se.fit)
hi_r  <- invlogit(pr$fit + 1.96 * pr$se.fit)

# plot CI segments only, labeled by the original liner names
op <- par(mar = c(5, 10, 4, 2))

plot(
  NULL,
  xlim = range(c(lo_r, hi_r)),
  ylim = c(0.5, length(levs) + 0.5),
  yaxt = "n",
  xlab = "Predicted infected proportion (95% CI)",
  ylab = "",
  main = "Predicted mean infection by liner"
)

segments(lo_r, seq_along(levs), hi_r, seq_along(levs), lwd = 3)

axis(2, at = seq_along(levs), labels = levs, las = 1, cex.axis = 0.75)

par(op)
```


<br>

## Other named distributions

- Non-normal regression is increasingly popular!

    - Gamma, cauchy, laplace, etc.
    
- Non-normal distributions are everywhere

    - It's weirder to see a normal distribution than a non-normal
    
- Why do we use the normal then?

- When do we use non-normal regression?

<br>














