# Day 21

## Announcements

- Congratulations to The Outliers and Abra-ca-data!

    - Very positive feedback on all of you

    - Remember: participation is a resume bullet (see: "Hackathon")
    
    - Sign up for next year and bring your friends!
    
- Takeaway from datafest: keep it simple, stupid

    - Clearly communicable results with large, readable figures
    
    - Keep that in mind for your final project
    
- Today and tomorrow

    - Networks and graphs
    
    - Random effects
    
- This week: model fitting

    - Due Saturday, April 18 

    - Last call to struggle with R coding
    
- Next week: Results

    - Decide on the figures/tables you'd like to use
    
    - Discuss the outcomes of fitting your models
    
    - Put it all in the context of the data / science
    
- Apr 26 - May 2: Early presentations and rough drafts

- May 3 - May 12: Presentations

- May 15: Project submission deadline

- Portfolio

    - Compile all of the assignments you've done for the course as PDFs
    
    - Combine them into one document with a cover sheet
    
    - Extra credit
    
    - "Do you have data analytics / R programming experience?"
    
<br>

<br>

## Review

>All models are wrong, but some are very useful - George Box

When we develop a statistical model, we seek to represent the data generating process

<br>

- Averages, sums, aggregates

$$
y_i \sim N(\mu, \sigma^2)
$$

- Normal distributions aren't normal

    - They're very useful
    
```{r echo=FALSE, message=FALSE, warning=FALSE}
cols <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd")
x <- seq(-6, 6, length.out = 500)
plot(x, dnorm(x, 0, 1), type = "l", col = cols[1], lwd = 2,
     ylim = c(0, 0.9), xlab = "x", ylab = "Density",
     main = "Normal Distribution")
lines(x, dnorm(x, 0, 0.5),  col = cols[2], lwd = 2)
lines(x, dnorm(x, 0, 2),    col = cols[3], lwd = 2)
lines(x, dnorm(x, -2, 1),   col = cols[4], lwd = 2)
lines(x, dnorm(x, 2, 1.5),  col = cols[5], lwd = 2)
legend("topright", bty = "n", lwd = 2, col = cols, cex = 0.7,
       legend = c("mu=0,sig=1", "mu=0,sig=0.5", "mu=0,sig=2",
                  "mu=-2,sig=1", "mu=2,sig=1.5"))
```

    
<br>

- Infections, sightings, successes

$$
y_i \sim \text{Bern}(p)
$$

- If we sum many bernoulli r.v.s:

$$
\sum_{i=1}^N y_i \sim \text{Bin}(N,p)
$$


<br>

- Counts, occurrences

$$
y_i \sim \text{Pois}(\lambda)
$$

- There are many times poisson r.v.s will look binomial

    - Common sense usually points us to the right one
    
```{r echo=FALSE, message=FALSE, warning=FALSE}
lambdas <- c(1, 2, 5, 10, 20)
x_pts   <- 0:35
plot(NULL, xlim = c(0, 35), ylim = c(0, 0.40),
     xlab = "k", ylab = "P(X = k)", main = "Poisson Distribution")
for (i in seq_along(lambdas)) {
  points(x_pts, dpois(x_pts, lambdas[i]), type = "h",
         col = cols[i], lwd = 2)
  points(x_pts, dpois(x_pts, lambdas[i]), pch = 16,
         col = cols[i], cex = 0.6)
}
legend("topright", bty = "n", lwd = 2, col = cols, cex = 0.7,
       legend = paste0("λ=", lambdas))

```

    
<br>

- Probabilities, proportions, rates

$$
y_i \sim \text{Beta}(\alpha, \beta)
$$

```{r echo=FALSE, message=FALSE, warning=FALSE}
x <- seq(0.001, 0.999, length.out = 500)
ab <- list(c(0.5,0.5), c(1,1), c(2,5), c(5,2), c(2,2))
ylim_beta <- c(0, 3)
plot(x, dbeta(x, ab[[1]][1], ab[[1]][2]), type = "l",
     col = cols[1], lwd = 2, ylim = ylim_beta,
     xlab = "x", ylab = "Density", main = "Beta Distribution")
for (i in 2:length(ab))
  lines(x, dbeta(x, ab[[i]][1], ab[[i]][2]), col = cols[i], lwd = 2)
legend("top", bty = "n", lwd = 2, col = cols, cex = 0.7,
       legend = c("a=0.5,b=0.5","a=1,b=1","a=2,b=5",
                  "a=5,b=2","a=2,b=2"))
```


- Special cases:

```{r}
hist(rbeta(10000,1,1))
```

```{r}
hist(runif(10000))
```

<br>

- Crop yields, birthing weights

$$
y_i \sim \text{Gamma}(\alpha,\beta)
$$

- Useful method for modeling non-constant (mean dependent) variance

```{r echo=FALSE, message=FALSE, warning=FALSE}
x <- seq(0.001, 20, length.out = 500)

ag <- list(c(1,1), c(2,1), c(3,1), c(5,1), c(2,0.5))
plot(x, dgamma(x, ag[[1]][1], rate = ag[[1]][2]), type = "l",
     col = cols[1], lwd = 2, ylim = c(0, 1),
     xlab = "x", ylab = "Density", main = "Gamma Distribution")
for (i in 2:length(ag))
  lines(x, dgamma(x, ag[[i]][1], rate = ag[[i]][2]),
        col = cols[i], lwd = 2)
legend("topright", bty = "n", lwd = 2, col = cols, cex = 0.7,
       legend = c("a=1,b=1","a=2,b=1","a=3,b=1",
                  "a=5,b=1","a=2,b=2"))
```


<br>

Honorable mentions:

- Laplace / Double exponential

    - Useful when wanting to adjust the linear model for absolute loss instead of squared loss

```{r echo=FALSE, message=FALSE, warning=FALSE}
dlaplace <- function(x, mu = 0, b = 1)
  (1 / (2 * b)) * exp(-abs(x - mu) / b)

x <- seq(-8, 8, length.out = 500)
params_l <- list(c(0,1), c(0,0.5), c(0,2), c(-2,1), c(2,1.5))
plot(x, dlaplace(x, params_l[[1]][1], params_l[[1]][2]),
     type = "l", col = cols[1], lwd = 2, ylim = c(0, 1),
     xlab = "x", ylab = "Density", main = "Laplace Distribution")
for (i in 2:length(params_l))
  lines(x, dlaplace(x, params_l[[i]][1], params_l[[i]][2]),
        col = cols[i], lwd = 2)
legend("topright", bty = "n", lwd = 2, col = cols, cex = 0.7,
       legend = c("mu=0,b=1","mu=0,b=0.5","mu=0,b=2",
                  "mu=-2,b=1","mu=2,b=1.5"))
```

<br>

- Pareto distribution

    - 80/20 law: $80\%$ of outcomes are the result of $20\%$ of causes

```{r echo=FALSE, message=FALSE, warning=FALSE}
dpareto <- function(x, alpha, x_m = 1)
  ifelse(x < x_m, 0, alpha * x_m^alpha / x^(alpha + 1))

x <- seq(1, 8, length.out = 500)
alphas_p <- c(1, 1.5, 2, 3, 4)
y_all_p  <- sapply(alphas_p, function(a) dpareto(x, a))
plot(x, y_all_p[,1], type = "l", col = cols[1], lwd = 2,
     ylim = c(0, 4), xlim = c(1,5),
     xlab = "x", ylab = "Density", main = "Pareto Distribution")
for (i in 2:length(alphas_p))
  lines(x, y_all_p[,i], col = cols[i], lwd = 2)
legend("topright", bty = "n", lwd = 2, col = cols, cex = 0.7,
       legend = paste0("a=", alphas_p, ", xm=1"))
```

<br>

- Power law

    - Pareto is just one type of power law distribution

```{r echo=FALSE, message=FALSE, warning=FALSE}
dpowerlaw <- function(x, alpha, x_min = 1)
  ifelse(x < x_min, 0, (alpha - 1) / x_min * (x / x_min)^(-alpha))

x <- seq(1, 10, length.out = 500)
alphas <- c(2, 2.5, 3, 3.5, 4)
y_all  <- sapply(alphas, function(a) dpowerlaw(x, a))
plot(x, y_all[,1], type = "l", col = cols[1], lwd = 2,
     ylim = c(0, max(y_all)), xlim = c(1,5),
     xlab = "x", ylab = "Density", main = "Power Law")
for (i in 2:length(alphas))
  lines(x, y_all[,i], col = cols[i], lwd = 2)
legend("topright", bty = "n", lwd = 2, col = cols, cex = 0.7,
       legend = paste0("a=", alphas))
```

<br>

Generalized linear models

- Simply change the distributional support of the linear model

    - Normal regression / beta regression / binomial regression...

    - *Insert distribution* regression

- Pros: adjustment of assumption

    - Non-normal data can break a lot of assumptions at once beyond just normality
    
    - Adjusting for that can get a lot of inferential and predictive power back
    
- Cons: loss of classical inference

    - We're constantly improve inferential techniques for GLMs
    
    - Remember that an F-test relies on normality at it's core
    
    - t-test relies on *asymptotic* normality
    
    - This trend continues
    
<br>
  
<br>
  
**Any questions?**
  
<br>
  
  
## Networks and Graphs
  
- Every math based science has a foundational math field that built it
  
- For statistics that's probability theory
  
- All of network sciences begins with graph theory
  
- Graph theory begins with Euler and the [Seven Bridges of Konigsberg](https://en.wikipedia.org/wiki/Seven_Bridges_of_K%C3%B6nigsberg)
  
<br>

### Definitions and notation

A graph is defined as $G = (V, E)$ where:

- $V$ is the **vertex set**

    - A vertex is an individual

- $E$ is the **edge set**

    - An edge is a connection between individuals
    
<br>
    
A graph is considered **undirected** if $G_{ij} = G_{ji}$

- a.k.a. the direction of edge conntection is irrelevant

    - We'll be focusing on undirected graphs today
    
    - Directed graphs are more common in causal studies

<br>

An **adjacency matrix** ($A$) is the matrix representation of a graph

- Vertices are represented as rows and columns ($i,j$)

- If an edge exists between $i$ and $j$ then $A_{i,j} = A_{j,i} =  1$

- If there is no edge between $i$ and $j$ then $A_{i,j} = A_{j,i} = 0$

<br>

- (WhiteboarD) Basic graph examples

    - Fully connected
    
    - Clusters
    
    - Small world

<br>

### Building a network

```{r adjacency-matrix, message=FALSE, warning=FALSE}
library(tidyverse)

n = 10

# node communitys
community = c(1, 1, 1, 2, 2, 2, 3, 3, 3, 3)
node_labels = paste0("V", 1:n)

# empty adjacency matrix
A = matrix(0, nrow = n, ncol = n,
           dimnames = list(node_labels, node_labels))

# function to add edges
add_edge = function(A, i, j) { A[i, j] <- 1; A[j, i] <- 1; A }

# community edges
# community A
A = add_edge(A, 1, 2); A = add_edge(A, 1, 3); A = add_edge(A, 2, 3)
# community B
A = add_edge(A, 4, 5); A = add_edge(A, 4, 6); A = add_edge(A, 5, 6)
# community C
A = add_edge(A, 7, 8); A <- add_edge(A, 7, 9); A <- add_edge(A, 8, 9)
A = add_edge(A, 8, 10); A <- add_edge(A, 9, 10)

# cross-community edges
A = add_edge(A, 3, 4)   # A <-> B (node 3 and 4)
A = add_edge(A, 6, 7)   # B <-> C (node 6 and 7)

print(A)
```

```{r}
# undirected implies symmetric
all(A == t(A))

# E = sum(Adjacency Matrix) / 2 (since it's symmetric)
sum(A) / 2
```

<br>

### Basic analysis

The **degree** $k_i$ of vertex $i$ is the sum of each row in $A$:
$$k_i = \sum_{j} A_{ij}$$

```{r}
degree = rowSums(A)
degree

# degree distribution visualized
barplot(degree, col = "white")
```

<br>

The **shortest path** refers to the path from $i$ to $j$ that crosses the least edges of all possible paths.

- **Betweenness** quantifies how often a node lies on the shortest path between two other nodes

<br>

To find shortest paths we use a shortest path **algorithm** to search the network.

- There's a lot of diversity for algorithms (Dijkstra / Bellman-Ford / A-star)

- Since the network is small we'll just use a **breadth-first search**

    - Algorithm logic on the whiteboard
    
- Note: I'm implementing these algorithms from scratch

    - Extremely impractical, I just don't want to use packages for this lecture
    
    - In practice you can use the `igraph` package in R or NetworkX library in python
    
<br>

```{r}
# standard breadth-first search algorithm
bfsd = function(A, source) {
  n = nrow(A)
  dist = rep(Inf, n)
  dist[source] = 0
  q = source
  while (length(q) > 0) {
    u = q[1]
    q = q[-1]
    neighbors = which(A[u, ] == 1)
    for (v in neighbors) {
      if (is.infinite(dist[v])) {
        dist[v] = dist[u] + 1
        q = c(q, v)
      }
    }
  }
  dist
}
```

```{r}
# brandes algorithm for betweenness
# implementing bfs as the sp algorithm
betweenness = function(A) {
  n = nrow(A)
  cb = rep(0, n)

  for (s in 1:n) {
    dist = rep(-1, n); dist[s] = 0
    sigma = rep(0, n);  sigma[s] = 1
    pred = vector("list", n)
    q = s
    order_list = c()

    while (length(q) > 0) {
      v = q[1]; q = q[-1]
      order_list = c(order_list, v)
      for (w in which(A[v, ] == 1)) {
        if (dist[w] < 0) {
          dist[w] = dist[v] + 1
          q = c(q, w)
        }
        if (dist[w] == dist[v] + 1) {
          sigma[w] = sigma[w] + sigma[v]
          pred[[w]] = c(pred[[w]], v)
        }
      }
    }

    delta = rep(0, n)
    for (w in rev(order_list)) {
      for (v in pred[[w]]) {
        delta[v] = delta[v] + (sigma[v] / sigma[w]) * (1 + delta[w])
      }
      if (w != s) cb[w] = cb[w] + delta[w]
    }
  }
  cb / 2
}

btwn = betweenness(A)
names(btwn) = node_labels
btwn

# betweennes distribution
# high betweenness is expected at "bridge" vertices
barplot(btwn, col = "white")
```

<br>

- The eigenvector centrality tells us about the *influence* of certain vertices

    - Solve for the leading eigenvector of $A$:

```{r eigenvector-centrality}
ev = eigen(A)
ev_cent = abs(ev$vectors[, 1])
ev_cent = ev_cent / max(ev_cent) # normalize to [0,1]
names(ev_cent) = node_labels

round(ev_cent, 3)
```

<br>

### Community detection

Note: there are many method for community detection but since this is a stats class we'll stick to statistical pattern recognition.

- Consider two graphs, $G_0$ and $G_A$

    - $G_0$ arises from a uniform, random process
    
    - If a graph, $G$, arises from the same process as $G_0$, it should have no communities
    
    - $G_A$ arises from a process where communities form
    
- We can measure the degree with which $G \in G_0$ using **modularity**

$$Q = \frac{1}{2m} \sum_{ij} \left[ A_{ij} - \frac{k_i k_j}{2m} \right] \delta(c_i, c_j)$$

Where $m$ is the total number of edges, $k_i$ is the degree of vertex $i$, and $\delta(c_i, c_j) = 1$ if vertex $i$ and $j$ are in the same community.

```{r}
compute_modularity = function(A, community) {
  m = sum(A) / 2
  k = rowSums(A)
  n = nrow(A)
  Q = 0
  for (i in 1:n) {
    for (j in 1:n) {
      if (community[i] == community[j]) {
        Q = Q + (A[i, j] - (k[i] * k[j]) / (2 * m))
      }
    }
  }
  Q / (2 * m)
}

Q_true = compute_modularity(A, community)
as.numeric(Q_true)
```

- This value implies there's communities present and our graph does not belong to $G_0$

    - From here we can use search algorithms to define communities from vertex features
    
    - There are also detection algorithms (i.e., greedy search) that focus on pure community classification
    
<br>

### Erdos–Renyi

An **Erdos–Renyi (ER) graph** $G(n, p)$ is generated by taking $n$ vertices and including each possible edge independently with probability $p$.

$$\mathbb{E}[m] = p \binom{n}{2}$$

- The degree distribution of an Edros-Reyni vertex is binomial

    - As $n \rightarrow \infty$ it becomes poisson
    
<br>

```{r}
erdos_reyni = function(n, p) {
  A = matrix(0, nrow = n, ncol = n)
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      if (runif(1) < p) {
        A[i, j] = 1
        A[j, i] = 1
      }
    }
  }
  rownames(A) = colnames(A) = paste0("V", 1:n)
  A
}

set.seed(73)
ER = erdos_reyni(n = 10, p = 0.3)
ER
```

```{r}
barplot(rowSums(ER), col = "white")
```


<br>

### Why graphs?

- Consider the scenario of disease spread

    - Where you are and what time you're there are clearly important
    
    - But what about contact based disease? (i.e., COVID, Mono, STDs)
    
    - Then it also matters *who* you interacted with
    
<br>

- In a situation where the graph arises randomly:

    - There's no real need to understand network features
    
    - No knowledge of network structure is equivalent to knowledge of a random structure
    
- If a graph arises from a stricter process?

    - Individuals who have a lot of contact will be the most threatening to the total graph
    
    - If we identify them they contain most of our information
    
    - (Whiteboard) Cheerleader / Football player model 
    
<br>

- Networks are more concerned with modeling expectations and rate

    - What will this trend towards?
    
- The goal is to deflate varince by uncovering more of the network's structure

<br>

### Simulation study

We can simulate an outbreak across a network with a "Susceptible - Infected" (SI) model

- Assumes that an individual is in one of two states

    - Able to be infected or actively infected

- Typical expansion: "Susceptible - Infected - Recovered" (SIR) model

    - We'll avoid this to keep things simple but this is closer to reality since recovery tends to imply temporary immunity
    
- SI is a great model for non-lethal pathogens like HSV where contact is required for infection

    - Vertex 1 is infected
    
    - Vertex 1 connects with Vertex 2
    
    - Vertex 2 is infected
    
    - Those two share no further connections
    
    - The outbreak is cut off at Vertex 2
    
<br>

```{r}
si_simulation = function(A, patient_zero = 1, beta = 0.4, max_t = 20) {
  n = nrow(A)
  status = rep("S", n)
  status[patient_zero] = "I"
  inf_time = rep(NA_real_, n)
  inf_time[patient_zero] = 0

  for (t in 1:max_t) {
    new_status = status
    for (j in which(status == "S")) {
      infected_neighbors = sum(A[j, ] == 1 & status == "I")
      if (infected_neighbors > 0) {
        p_inf = 1 - (1 - beta)^infected_neighbors
        if (runif(1) < p_inf) {
          new_status[j] = "I"
          inf_time[j] = t
        }
      }
    }
    status = new_status
    if (all(status == "I")) break
  }
  inf_time
}
```

```{r}
n_sims = 100
beta = 0.4
patient_zero = 1
max_t = 20
t_grid = 0:max_t

curves = function(adj_matrix, n_sims, beta, patient_zero, max_t) {
  curves = matrix(0, nrow = n_sims, ncol = max_t + 1)
  inf_times_all = matrix(NA, nrow = n_sims, ncol = nrow(adj_matrix))
  for (s in 1:n_sims) {
    inf_t = si_simulation(adj_matrix, patient_zero, beta, max_t)
    inf_times_all[s, ] = inf_t
    for (t in t_grid) {
      curves[s, t + 1] = mean(!is.na(inf_t) & inf_t <= t)
    }
  }
  list(curves = curves, inf_times = inf_times_all)
}

res_struct = curves(A, n_sims, beta, patient_zero, max_t)

er_curves = matrix(0, nrow = n_sims, ncol = max_t + 1)
er_inf_times = matrix(NA, nrow = n_sims, ncol = n)
for (s in 1:n_sims) {
  A_er_s = erdos_reyni(n, 0.3)
  inf_t = si_simulation(A_er_s, patient_zero, beta, max_t)
  er_inf_times[s, ] = inf_t
  for (t in t_grid) {
    er_curves[s, t + 1] = mean(!is.na(inf_t) & inf_t <= t)
  }
}
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
mean_struct = colMeans(res_struct$curves)
sd_struct = apply(res_struct$curves, 2, sd)
mean_er = colMeans(er_curves)
sd_er = apply(er_curves, 2, sd)

plot(t_grid, mean_struct,
     type = "l", lwd = 2, col = "#E07B54",
     ylim = c(0, 1),
     xlab = "Time Step",
     ylab = "Proportion Infected",
     main = "",
     las = 1)

polygon(c(t_grid, rev(t_grid)),
        c(pmin(mean_struct + sd_struct, 1), rev(pmax(mean_struct - sd_struct, 0))),
        col = adjustcolor("#E07B54", alpha.f = 0.2), border = NA)

lines(t_grid, mean_er, lwd = 2, col = "#5B8DB8")
polygon(c(t_grid, rev(t_grid)),
        c(pmin(mean_er + sd_er, 1), rev(pmax(mean_er - sd_er, 0))),
        col = adjustcolor("#5B8DB8", alpha.f = 0.2), border = NA)

legend("bottomright",
       legend = c("Original Graph", "Erdos–Renyi"),
       col    = c("#E07B54", "#5B8DB8"),
       lwd    = 2, bty = "n")
abline(h = c(0.5, 1), lty = 3, col = "gray60")
```

<br>

- An observation

```{r}
# similar edge count
sum(A) / 2
sum(ER) / 2
```

```{r}
# similar average degree
mean(rowSums(A))
mean(rowSums(ER))
```

- Why does the infection curve differ for these two graphs?

<br>

<br>

```{r}
A # original graph
ER # erdos-reyni graph
```

- Community structures inform decision making

    - Measures like betweenness can identify points we can 'sever' vulnerabilities
    
```{r}
barplot(btwn, col = "white")
```

<br>


```{r}
closeness = function(A) {
  n = nrow(A)
  cc = numeric(n)
  for (i in 1:n) {
    d = bfsd(A, i)
    reachable = d[is.finite(d) & d > 0]
    cc[i] = if (length(reachable) > 0) 1 / mean(reachable) else 0
  }
  cc
}

clsns = closeness(A)
names(clsns) = node_labels

early_cutoff = 5
outcome_mat = (res_struct$inf_times <= early_cutoff)

sim_df = do.call(rbind, lapply(1:n_sims, function(s) {
  data.frame(
    node = 2:n,
    sim = s,
    early_inf = as.integer(outcome_mat[s, 2:n] & !is.na(res_struct$inf_times[s, 2:n])),
    degree = degree[2:n],
    closeness = clsns[2:n],
    betweenness = btwn[2:n],
    community = community[2:n]
  )
}))

sim_df$early_inf[is.na(sim_df$early_inf)] = 0

cat(sprintf("Early infection rate: %.1f%%\n", 100 * mean(sim_df$early_inf)))
```

<br>

- We can use network metrics in statistical model development

    - We'll find they're powerful predictors when we're "classifying"

```{r}
m1 = glm(early_inf ~ scale(degree),
          data = sim_df,
          family = binomial(link = "logit"))

m2 = glm(early_inf ~ scale(closeness),
          data = sim_df,
          family = binomial(link = "logit"))

m3 = glm(early_inf ~ scale(degree) + scale(closeness) + scale(betweenness),
          data = sim_df,
          family = binomial(link = "logit"))

AIC(m1,m2,m3)

summary(m3)
```

<br>


### k-Nearest-Neighbors

Data can be analyzed strictly through it's geometry

- One of those methods is known as "k-Nearest-Neighbors"

- We assume that things tend to be similar to things they're close to

    - The $k$ nearest neighbors of a given vertex are checked for a classification
    
    - The majority class of those neighbors is the class assigned to the vertex
    
- This is a form of non-parametric statistics where we "train" on every variable

    - If there's some structure we can isolate from those variable relationships, it shows itself with knn

```{r}
library(MASS)
data(Pima.tr)

set.seed(42)
# shuffle and split 
idx = sample(nrow(Pima.tr))
pima = Pima.tr[idx, ]
split = floor(0.8 * nrow(pima))
train_df = pima[1:split, ]
test_df = pima[(split + 1):nrow(pima), ]

# set up data partitions
train_means = colMeans(train_df[, 1:7])
train_sds = apply(train_df[, 1:7], 2, sd)
scale_data = function(X, means, sds) {
  sweep(sweep(X, 2, means, "-"), 2, sds, "/")
}
train_X = scale_data(as.matrix(train_df[, 1:7]), train_means, train_sds)
test_X = scale_data(as.matrix(test_df[, 1:7]),  train_means, train_sds)
train_y = as.character(train_df$type)
test_y = as.character(test_df$type)

# knn algorithm
knn = function(train_X, train_y, test_X, k) {
  apply(test_X, 1, function(query) {
    dists = sqrt(rowSums((train_X - query)^2))
    nn_idx = order(dists)[1:k]
    nn_labs = train_y[nn_idx]
    names(sort(table(nn_labs), decreasing = TRUE))[1]
  })
}

# knn predictions
k_full = nrow(train_X)
pred_k3 = knn(train_X, train_y, test_X, k = 3)
pred_kn = knn(train_X, train_y, test_X, k = k_full)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# confusion matrices
cat("\nConfusion Matrix (k = 3):\n")
print(table(Predicted = pred_k3, Actual = test_y))

cat("\nConfusion Matrix (k = n):\n")
print(table(Predicted = pred_kn, Actual = test_y))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# left plot - k = 3, right plot - k = n
diag_cols = c("No" = "#5B8DB8", "Yes" = "#E07B54")
par(mfrow = c(1,2), mar = c(4.5,4,1,1))
for (run in list(list(k = 3, pred = pred_k3),
                 list(k = k_full, pred = pred_kn))) {

  correct = run$pred == test_y
  label = if (run$k == 3) "k = 3" else sprintf("k = n  (%d)", k_full)

  plot(test_df$glu, test_df$bmi,
       col = diag_cols[test_y],
       pch = ifelse(correct, 19, 4),
       cex = 1.6,
       xlab = "Plasma Glucose",
       ylab = "BMI")

  legend("topleft",
         legend = c("No diabetes", "Diabetes", "Correct", "Misclass"),
         col = c(diag_cols, "black", "black"),
         pch = c(19, 19, 19, 4),
         bty = "n", cex = 0.8)
}
```

- The primary reasons we like this method:

    - "Assumption free"
    
    - Supervised learning with minimal guidance
    
    - Easily scaled
    
- Assumptions can quickly make this a better method

    - We don't know how long every assumption will hold
    
    - Assumption: Humans will get diabetes (probably solid)
    
    - Assumption: Diabetes relates to BMI (shakey, BMI is imperfect)

<br>

### Further discussion

- Directed acyclic graphs (DAGs)

    - Causal relationships
    
    - Causal search algorithms, variable rejection, rationalizing confounders

- Bayes / Belief networks

    - Dependence graphs

    - Statistical machine learning 

- Markov random fields

    - Markov properties / processes
    
    - Sparse matrices versus dense matrices
    
<br>

## Go away
