# Day 22

## Effects

Assumptions of statistics

- Each observation

    - Independent
    
    - Identically distributed

- A study is conducted at a research hospital

    - The subjects share location, nursing staff, equipment, environmental conditions
    
- Our subjects are more similar to each other than to outsiders who share their same characteristics
    
- We don't measure the background variables that dictate this phenomenon

- **Psuedoreplication**

    - General linear model *assumes* i.i.d. observations
    
    - Skews confidence bands and p-values

<br>

```{r pooling-illustration, echo=FALSE, fig.width=9, fig.height=8}

set.seed(7)
n_groups = 5
n_per    = 12
group    = rep(1:n_groups, each = n_per)
x        = rnorm(n_groups * n_per)

group_int_fx   = c(-4, -1.5, 0, 1.5, 3.5)
group_slope_fx = c(-0.8, -0.3, 0, 0.4, 0.9)

y = 2 + (0.8 + group_slope_fx[group]) * x +
     group_int_fx[group] +
     rnorm(n_groups * n_per, sd = 1)

group_cols = c("#E07B54", "#5B8DB8", "#6AAB6A", "#9B7EC8", "#C4A35A")

xlim = range(x)
ylim = range(y)

global_int   = mean(tapply(y - 0.8 * x, group, mean))
group_ints   = tapply(y - 0.8 * x, group, mean)
shrunk_ints  = 0.6 * group_ints + 0.4 * global_int

global_slope  = coef(lm(y ~ x))[2]
group_slopes  = sapply(1:n_groups, function(g) coef(lm(y[group==g] ~ x[group==g]))[2])
shrunk_slopes = 0.6 * group_slopes + 0.4 * global_slope

par(mfrow = c(2, 2), mar = c(4, 4, 3.5, 1))

plot(x, y, col = adjustcolor(group_cols[group], 0.5), pch = 19, cex = 0.9,
     main = "Fixed Intercept, Fixed Slope",
     xlab = "x", ylab = "y", xlim = xlim, ylim = ylim)
abline(lm(y ~ x), lwd = 2.5, col = "black")

plot(x, y, col = adjustcolor(group_cols[group], 0.5), pch = 19, cex = 0.9,
     main = "Random Intercepts, Fixed Slope",
     xlab = "x", ylab = "y", xlim = xlim, ylim = ylim)
for (g in 1:n_groups) {
  abline(a = shrunk_ints[g], b = 0.8, col = group_cols[g], lwd = 2)
}

plot(x, y, col = adjustcolor(group_cols[group], 0.5), pch = 19, cex = 0.9,
     main = "Fixed Intercept, Random Slopes",
     xlab = "x", ylab = "y", xlim = xlim, ylim = ylim)
for (g in 1:n_groups) {
  abline(a = global_int, b = shrunk_slopes[g], col = group_cols[g], lwd = 2)
}

plot(x, y, col = adjustcolor(group_cols[group], 0.5), pch = 19, cex = 0.9,
     main = "Random Intercepts, Random Slopes",
     xlab = "x", ylab = "y", xlim = xlim, ylim = ylim)
for (g in 1:n_groups) {
  abline(a = shrunk_ints[g], b = shrunk_slopes[g], col = group_cols[g], lwd = 2)
}

par(mfrow = c(1, 1))
```

<br>

- Choice of random versus fixed effect is very subjective (don't sue me)

    - Our goal today will be to test the idea of what to assign as random

<br>

<br>

## Mixed models

South African Heart Disease

- Retrospective sample of 462 males from a high-risk region of the Western
Cape collected by Rossouw et al. (1983).

```{r load-saheart}
url = "https://hastie.su.domains/ElemStatLearn/datasets/SAheart.data"
SAheart = read.csv(url)
head(SAheart)
```

<br>

- Age can be a pain to extract inference from

    - We need everything inbetween the min and max
    
    - Otherwise it gets messy
    
- Decile groups can help with that problem

```{r create-age-groups}
library(nlme)
library(tidyverse)

SAheart$age_group = cut(SAheart$age,
                         breaks = quantile(SAheart$age,
                                           probs = seq(0, 1, by = 0.2)),
                         include.lowest = TRUE,
                         labels   = paste0("D", 1:5))


print(table(SAheart$age_group))
print(tapply(SAheart$age, SAheart$age_group, range))
```


```{r cardio-viz, echo=FALSE}
group_cols = c("#E07B54", "#5B8DB8", "#6AAB6A", "#9B7EC8", "#C4A35A")
names(group_cols) = paste0("D", 1:5)

par(mar = c(4, 3, 3, 1))
for (g in paste0("D", 1:5)) {
  sub = SAheart[SAheart$age_group == g, ]
  plot(sub$ldl, sub$sbp,
       col  = group_cols[g],
       pch  = 19, cex = 0.8,
       xlab = "LDL", ylab = "SBP",
       main = paste0("Age group ", g, "\n(n=", nrow(sub), ")"),
       ylim = range(SAheart$sbp),
       xlim = range(SAheart$ldl))
  abline(lm(sbp ~ ldl, data = sub), col = "black", lwd = 1.5)
}
par(mfrow = c(1, 1))
```

```{r cardio-group-means, echo=FALSE}
group_means = tapply(SAheart$sbp, SAheart$age_group, mean)
barplot(group_means,
        col    = group_cols,
        border = "white",
        main   = "Mean Systolic Blood Pressure by Age Group",
        xlab   = "Age Decile Group",
        ylab   = "Mean SBP (mmHg)",
        ylim   = c(0, max(group_means) * 1.2))
abline(h = mean(SAheart$sbp), lty = 2, col = "gray50")
legend("topleft", legend = "Global mean", lty = 2, lwd = 2,
       col = "gray50", bty = "n")
```

- (Whiteboard) Writing out random effects

```{r cardio-models}
# fixed effects
m1 = lm(sbp ~ ldl + tobacco + obesity, data = SAheart)

# random intercepts
m2 = lme(sbp ~ ldl + tobacco + obesity,
                   random = ~ 1 | age_group,
                   data   = SAheart,
                   method = "ML")

# random slopes on ldl
m3 = lme(sbp ~ ldl + tobacco + obesity,
                     random = ~ 0 + ldl | age_group,
                     data   = SAheart,
                     method = "ML")

# random intercepts and slopes
m4 = lme(sbp ~ ldl + tobacco + obesity,
                    random = ~ 1 + ldl | age_group,
                    data   = SAheart,
                    method = "ML")

summary(m1)
summary(m2)

coef(m1)
fixef(m2)
fixef(m3)
fixef(m4)
```




```{r cardio-coef-compare, echo=FALSE, fig.width=9, fig.height=5}

preds = c("ldl", "tobacco", "obesity")

coef_fixed      = coef(summary(m1))[preds, ]
coef_rand_int   = summary(m2)$tTable[preds, ]
coef_rand_slope = summary(m3)$tTable[preds, ]
coef_rand_both  = summary(m4)$tTable[preds, ]

ests = list(
  coef_fixed[, "Estimate"],
  coef_rand_int[, "Value"],
  coef_rand_slope[, "Value"],
  coef_rand_both[, "Value"]
)

ses = list(
  coef_fixed[, "Std. Error"],
  coef_rand_int[, "Std.Error"],
  coef_rand_slope[, "Std.Error"],
  coef_rand_both[, "Std.Error"]
)

model_cols = c(
  "Fixed"         = "#E07B54",
  "Rand. Int."    = "#5B8DB8",
  "Rand. Slope"   = "#6AAB6A",
  "Rand. Int+Slp" = "#9B7EC8"
)

par(mfrow = c(1, 3), mar = c(6, 4, 3, 1))

for (i in seq_along(preds)) {
  est_i = sapply(ests, `[`, i)
  se_i  = sapply(ses,  `[`, i)
  ylim  = range(c(est_i - 2 * se_i, est_i + 2 * se_i))

  plot(1:4, est_i,
       pch  = 19, cex  = 1.8,
       col  = model_cols,
       ylim = ylim,
       xaxt = "n",
       xlab = "",
       ylab = "Estimate",
       main = preds[i])
  axis(1, at = 1:4, labels = names(model_cols), las = 2, cex.axis = 0.72)
  segments(1:4, est_i - 1.96 * se_i,
           1:4, est_i + 1.96 * se_i,
           col = model_cols, lwd = 2)
  abline(h = 0, lty = 2, col = "gray60")
}
```

- `nlme::BodyWeight`

    - Longitudinal experiment tracking the body weight (grams) of **64 rats** measured over 64 days under three different diets (Hand & Crowder, 1996)

- Bioaccumulation

    - We have some baseline level of environmental toxicity (i.e., microplastics in water)
    
    - Wouldn't it make sense to account for that?


```{r load-bodyweight}
data(BodyWeight, package = "nlme")
head(BodyWeight)
```

```{r bw-viz, echo=FALSE}
diet_cols = c("1" = "#E07B54", "2" = "#5B8DB8", "3" = "#6AAB6A")

plot(BodyWeight$Time, BodyWeight$weight,
     col  = adjustcolor(diet_cols[BodyWeight$Diet], 0.3),
     pch  = 19, cex = 0.6,
     xlab = "Day",
     ylab = "Body Weight (g)",
     main = "Rat Body Weight Over Time by Diet Group")

for (rat in unique(BodyWeight$Rat)) {
  sub = BodyWeight[BodyWeight$Rat == rat, ]
  lines(sub$Time, sub$weight,
        col = adjustcolor(diet_cols[sub$Diet[1]], 0.4), lwd = 0.8)
}

legend("topleft",
       legend = paste("Diet", 1:3),
       col    = diet_cols,
       lwd    = 2, bty = "n")
```

```{r bw-boxplot, echo=FALSE}
boxplot(weight ~ Diet, data = BodyWeight,
        col    = diet_cols,
        border = "gray30",
        main   = "Body Weight Distribution by Diet Group",
        xlab   = "Diet",
        ylab   = "Weight (g)")
```



```{r bw-models}
# fixed intercepts fixed slopes
m1 = lm(weight ~ Time, data = BodyWeight)

# random intercepts fixed slopes
m2 = lme(weight ~ Time,
           random = ~ 1 | Diet,
           data   = BodyWeight,
           method = "ML")

# fixed intercept random slopes
m3 = lme(weight ~ Time,
           random = ~ 0 + Time | Diet,
           data   = BodyWeight,
           method = "ML")

# random intercepts random slopes
m4 = lme(weight ~ Time,
           random = ~ 1 + Time | Diet,
           data   = BodyWeight,
           method = "ML")

anova(m2, m3, m4)
```

```{r bw-fitted, echo=FALSE}
time_seq  = seq(min(BodyWeight$Time), max(BodyWeight$Time), length.out = 100)
diet_levs = levels(BodyWeight$Diet)

pred_m1 = predict(m1, newdata = data.frame(Time = time_seq))

pred_m2 = lapply(diet_levs, function(d) {
  predict(m2, newdata = data.frame(Time = time_seq,
                                   Diet = factor(d, levels = diet_levs)),
          level = 1)
})

pred_m3 = lapply(diet_levs, function(d) {
  predict(m3, newdata = data.frame(Time = time_seq,
                                   Diet = factor(d, levels = diet_levs)),
          level = 1)
})

pred_m4 = lapply(diet_levs, function(d) {
  predict(m4, newdata = data.frame(Time = time_seq,
                                   Diet = factor(d, levels = diet_levs)),
          level = 1)
})

model_preds = list(pred_m2, pred_m3, pred_m4)
model_names = c("Random Intercepts",
                 "Random Slopes",
                 "Random Int. + Slope")

par(mfrow = c(1, 4), mar = c(4, 4, 3, 1))

plot(BodyWeight$Time, BodyWeight$weight,
     col  = adjustcolor(diet_cols[BodyWeight$Diet], 0.25),
     pch  = 19, cex = 0.6,
     xlab = "Day", ylab = "Body Weight (g)",
     main = "Fixed Only")
lines(time_seq, pred_m1, lwd = 2.5, col = "black")

for (m in 1:3) {
  plot(BodyWeight$Time, BodyWeight$weight,
       col  = adjustcolor(diet_cols[BodyWeight$Diet], 0.25),
       pch  = 19, cex = 0.6,
       xlab = "Day", ylab = "Body Weight (g)",
       main = model_names[m])
  for (i in seq_along(diet_levs)) {
    lines(time_seq, model_preds[[m]][[i]],
          col = diet_cols[diet_levs[i]], lwd = 2.5)
  }
}
```

```{r bw-residuals, echo=FALSE}
par(mfrow = c(2,2), mar = c(4, 4, 3, 1))

resid_list  = list(residuals(m1), residuals(m2),
                    residuals(m3), residuals(m4))
resid_names = c("Fixed Only",       "Random Intercepts",
                 "Random Slopes",    "Random Int. + Slope")

for (i in 1:4) {
  boxplot(resid_list[[i]] ~ BodyWeight$Diet,
          col    = diet_cols,
          border = "gray30",
          main   = resid_names[i],
          xlab   = "Diet Group",
          ylab   = "Residual")
  abline(h = 0, lty = 2, col = "gray50")
}
```

## Assumption vs Data

- How do we improve prediction?

    - Add data
    
    - Add assumptions
    
- Assumptions are information

    - If all we have is assumptions, a model is the intersection of expert opinions
    
    - If all we have is data, a model is an exploration of the data
    
- Random forests / kNN 

    - Unsupervised learning for exploring data
    
    - Apply assumptions to the process and use the exploration to inform the final model
    
<br>

## Bayesian

- Originally everything was "Bayesian"

    - Inverse-probability
    
- Many of the results we seek in "frequentist" statistics

    - Imitations of their original bayesian counterparts
    
<br>

Why Bayes?

- Building a model for things we haven't observed (Rare / unheard of events)

- Capturing true processes with mechanistic models (presence / absence)

- Instant inferential output (some MLEs are made of duct tape and hope)

<br>

- Instead of MLE / MME / Kernel estimation / REML

    - Bayes theorem
    
- Once the model is made, it just works*

$$
P(A|B) = \frac{P(B|A)P(A)}{P(B|A)P(A)+P(B|A^c)P(A^c)}
$$

<br>

- *The denominator of Bayes theorem is a land mine

$$
P(\theta | y) = \frac{P(y | \theta) P(\theta)}{\int P(y | \theta) P(\theta) \, d\theta}
$$

<br>

- Let's develop a model

<br>

<br>

<br>

## Go away
